# Data transformation

**Learning objectives:**

- **Pick out rows** of a data frame with the **`dplyr::filter()`** function.
- **Sort rows** of a data frame with **`dplyr::arrange()`**.
- **Pick out columns** of a data frame with **`dplyr::select()`**.
- **Modify columns** of a data frame with **`dplyr::mutate()`**.
- **Group rows** of a data frame with **`dplyr::group()`**.
- **Apply functions to columns** of a (grouped) data frame with **`dplyr::summarize()`**.
- **Streamline data transformations** with the **pipe** operator (`%>%`).

## Introduction

### Prerequisites

dplyr is a package that provides functions to manipulate data frames. Data frame consists of columns (variables) and rows (observations).

dplyr is part of the tidyverse. You can install and load the all the packages from tidyverse (ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats)

```{r, eval=FALSE}
install.packages("tidyverse")
```

```{r}
library(tidyverse)
```

Or just install and load dplyr.

```{r, eval=FALSE}
install.packages("dplyr")
```

```{r}
library(dplyr)
```

### nycflights13

The data set `nycflights13` contains data about flights that departed New York City in 2013.

```{r, eval=FALSE}
install.packages("nycflights13")
```

```{r}
library(nycflights13)
```

When you use data set for the first time, it's good practice to quickly browse the data to check if you want to use the data.

To view the flights data in the R console.  .

```{r}
flights
```
To view the flights data in a spreadsheet-like viewer.

```{r, eval=FALSE}
View(flights)
```

Use `?flights` to open the help viewer to get info about all the variables.

```{r, eval=FALSE}
?flights
```

Check the size of the flights data frame using `nrow()`, `ncol()`, `length()`, and `dim()`.

```{r, eval=FALSE}
# number of rows
nrow(flights)
# [1] 336776

# number of columns
ncol(flights)
# [1] 19

# number of columns
length(flights)
# [1] 19

# number of rows and columns
dim(flights)
# [1] 336776     19
```

get the column names with `colnames()`

```{r}
colnames(flights)
```

## Comparisons and logical operators

### Comparisons

-   `>` greater than
-   `>=` greater than or equal
-   `<` less than
-   `<=` less than or equal
-   `==` equal
-   `!=` not equal


In math we use `=` for equality. In programming we use `==` for equality, and `=` for assignment.

compare numbers

```{r, eval=FALSE}
1 > 2
# [1] FALSE
1 >= 2
# [1] FALSE
1 < 2
# [1] TRUE
1 <= 2
# [1] TRUE
1 == 2
# [1] FALSE
1 != 2
# [1] TRUE
```

compare characters

```{r, eval=FALSE}
'a' > 'b'
# [1] FALSE
'a' >= 'b'
# [1] FALSE
'a' < 'b'
# [1] TRUE
'a' <= 'b'
# [1] TRUE
'a' == 'b'
# [1] FALSE
'a' != 'b'
# [1] TRUE
```

Case matters when comparing characters. For English, lowercase letters are less than uppercase letters.

```{r, eval=FALSE}
'a' > 'A'
# [1] FALSE
'a' >= 'A'
# [1] FALSE
'a' < 'A'
# [1] TRUE
'a' <= 'A'
# [1] TRUE
'a' == 'A'
# [1] FALSE
'a' != 'A'
# [1] TRUE
```

We can change case when comparing characters.

`tolower()` will change characters to lower case. `toupper()` will change characters to upper case.

```{r, eval=FALSE}
tolower('A')
# [1] "a"

toupper('a')
# [1] "A"
```


```{r, eval=FALSE}
'a' == tolower('A')
# [1] TRUE

toupper('a') == 'A'
# [1] TRUE
```

### Logical operators

`&` and; all expressions must be true in order to return true

```{r, eval=FALSE}
TRUE & TRUE
# [1] TRUE
TRUE & FALSE
# [1] FALSE
```

`|` or; one or more expressions must be true in order to return true; `|` is the key above the return key, not lowercase letter l.

```{r, eval=FALSE}
TRUE | TRUE
# [1] TRUE
TRUE | FALSE
# [1] TRUE
```

`!` not; negate the expression

```{r, eval=FALSE}
!TRUE
# [1] FALSE
!FALSE
# [1] TRUE
!!TRUE
# [1] TRUE
```

assign objects

```{r}
a <- 1
b <- 5
```

compare object and numbers

```{r, eval=FALSE}
a < 3
# [1] TRUE
a > 3
# [1] FALSE
a == 3
# [1] FALSE
a != 3
# [1] TRUE
```

compare objects

```{r, eval=FALSE}
a < b
# [1] TRUE
a > b
# [1] FALSE
a == b
# [1] FALSE
a != b
# [1] TRUE
```

comparison and logical operators

```{r, eval=FALSE}

a > 3
# [1] FALSE

b > 3
# [1] TRUE

a > 3 & b > 3
# [1] FALSE

a > 3 | b > 3
# [1] TRUE

!(a == b)
# [1] TRUE
```

## Filter rows with filter()

`filter()` allows you to pick out certain rows (observations) . `filter()` picks the rows which evaluates to TRUE for all criteria.

The first argument to `filter()` is a data frame, the subsequent arguments are the expressions.

Combine comparisons and logical operators on the columns to select rows.

```{r}
# number of rows and columns
dim(flights)
```

```{r}
# get the column names
colnames(flights)
```

```{r}
# select flights from November
filter(flights, month == 11)
```
`A tibble: 27,268 × 19` tells you 27,268 rows match the criteria


```{r}
# select flights from December
filter(flights, month == 12)
```

```{r}
# select flights not from December (e.g. January to November)
filter(flights, month != 12)
```

```{r}
# select flights from November or December
filter(flights, month == 11 | month == 12)
```
`&` (and)  vs `|` (or)

```{r}
# select flights from November or from 1st day of any month
filter(flights, month == 11 | day == 1)
```

```{r}
# select flights from November 1st
filter(flights, month == 11 & day == 1)
```

If you provide multiple comma-separated expressions, dplyr will automatically use `&` to combine the expressions.

```{r, eval=FALSE}
# both will select flights from November 1st

filter(flights, month == 11 & day == 1)

filter(flights, month == 11, day == 1)
```

dplyr functions do not change the original data. To save the results from a function, you need to assign the results to a object.

```{r, eval=FALSE}
nov1 <- filter(flights, month == 11, day == 1)

nrow(nov1)
## [1] 986

nrow(flights)
## [1] 336776

```

### Missing values

Often times, rows will not have data for certain columns. In spreadsheets, csv, tsv, the cells will be blank. In R data frames, the missing values are represented as `NA` (not available).

Operations with `NA` will return `NA`.

```{r, eval=FALSE}
NA > 5
# [1] NA

10 == NA
# [1] NA

NA + 10
# [1] NA

NA / 2
# [1] NA

NA == NA
# [1] NA
```

To check if value is missing use `is.na()`

```{r, eval=FALSE}
is.na(NA)
# [1] TRUE
```

By default, `filter()` excludes `NA` values. To include `NA` values, you must add an expression.

tibble is the tidyverse version of a data frame. tibble haves some extra functions that normal data frames do not have.

```{r}
# create a tibble with a column named "x", with 3 values
df <- tibble(x = c(1, NA, 3))
df
```


```{r}
# select rows with values greater than 1
filter(df, x > 1)
```

```{r}
# select rows with NA or values greater than 1
filter(df, is.na(x) | x > 1)
```

## Arrange rows with arrange()

`arrange()` changes the order of the rows. The first argument to `arrange()` is a data frame, the subsequent arguments are columns names are expression used to sort the rows.

Ascending (small to big) is the default order.

Use `desc()` for descending order (big to small).

```{r}
# sort flights by depature delay using ascending order
arrange(flights, dep_delay)
```

```{r}
# sort flights by depature delay using descending order.
arrange(flights, desc(dep_delay))
```

sort by multiple columns

```{r}
# sort flights by year, month, and day using ascending order
arrange(flights, year, month, day)
```

`arrange()` puts `NA` values at the end.

```{r}
# create tibble with 3 values
df <- tibble(x = c(1, NA, 3))

# sort puts NA at the end
arrange(df, x)
```

## Select columns with select()

`select()` lets you pick which columns (variables) to use. The first argument to `select()` is a data frame, the subsequent arguments are columns to use.

```{r}
# colnames() retrieves the column names
colnames(flights)
```

the order you list the columns will determine the order of the columns returned by `select()`.

```{r}
# select year, month, and day columns.
select(flights, year, month, day)
```

```{r}
# use ':' to select columns from year to day (inclusive).
select(flights, year:day)
```

```{r}
# use '-' to select columns except from year to day (inclusive).
select(flights, -(year:day))
```

helper functions for `select()`

`starts_with()`

```{r}
# select columns that start with "dep"
select(flights, starts_with("dep"))
```

`ends_with()`

```{r}
# select columns that end with "delay"
select(flights, ends_with("delay"))
```

`contains()`

```{r}
# select columns that contain  "dep"
select(flights, contains("dep"))
```

`matches()`

```{r}
# select columns that matches regular expression.
# "^a(.)r" means it starts with a, has any character, and then r.
select(flights, matches("^a(.)r"))
```

`num_range()`

```{r}
# create tibble with columns x1, x2, x3, x4
df <- tibble(x1 = c(1, 2), x2 = c(2, 3), x3 = c(4, 5), x4 = c(6, 7))

# select column that matches x1, x2 and x3
select(df, num_range("x", 1:3))
```

`rename()` changes the column names. `rename(data_frame, new_name = old_name)`

```{r}
# rename the column dep_time to departure_time

rename(flights, departure_time = dep_time)
```

use `select()` and `everything()` to move some columns to start of the dataframe

```{r}
# rearrange columns to time_hour, air_time, rest of the columns
select(flights, time_hour, air_time, everything())
```

## Add new variables with mutate()

`mutate()` adds new columns based on values from existing columns. Data frame includes existing and new columns.

```{r}
# create data frame with  7 columns: year, month, day, dep_delay, arr_delay, distance, air_time

flights_7_columns <- select(flights,
  year:day,
  ends_with("delay"),
  distance,
  air_time
)
```

```{r}
# calculate and add columns for gain, hours, and gain_per_hour

mutate(flights_7_columns,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

`transmute()` adds new columns based on values from existing columns. Data frame only includes new columns.

```{r}
# calculate and add columns for gain, hours, and gain_per_hour
transmute(flights_7_columns,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

### Useful creation functions

There are many functions to use with `mutate()`. The function must be vectorised: it must take a vector of values as input, return a vector with the same number of values as output.

Arithmetic operators: `+`, `-`, `*`, `/`, `^`

aggregate functions: `sum`, `mean`

Modular arithmetic: `%/%` (integer division) and `%%` (remainder)

Logs: `log()`, `log2()`, `log10()`

Offsets: `lead()` leading values; `lag()` lagging values.

Cumulative and rolling aggregates: `cumsum()` cumulative sums, `cumprod()` cumulative products, `cummin()` cumulative min, `cummax()` cumulative max, `cummean()` cumulative means.

Logical comparisons: `<`, `<=`, `>`, `>=`, `!=`, and `==`

Ranking: `min_rank(x)` gives ranks from smallest to largest. `min_rank(desc(x))` gives ranks from largest to smallest. `row_number()`, `dense_rank()`, `percent_rank()`, `cume_dist()`, `ntile()`.

Testing out various functions to see what they do.

```{r}
transmute(flights,
  dep_time,
  arr_time,
  # Modular arithmetic
  dep_hour = dep_time %/% 100,
  dep_minute = dep_time %% 100,
  # Arithmetic
  duration = arr_time - dep_time,
  # logs
  log_duration = log2(duration),
  # Offsets
  lead_duration = lead(duration)
)
```

```{r}
transmute(flights,
  dep_time,
  arr_time,
  # Arithmetic
  duration = arr_time - dep_time,
  # Cumulative aggregates
  cumsum_duration = cumsum(duration),
  # Logical comparisons
  long_duration = duration > 300,
  # Ranking
  rank_duration = min_rank(duration)
)
```

## Grouped summaries with summarize()

`summarize()` or `summarise()` collapses a data frame to a single row


By default, summarize includes NA values. `summarize()` will return `NA` if any values are `NA`.

```{r}
# calculate the mean departure delay for all the flights. Include NA.
summarize(flights, delay = mean(dep_delay))
```

Need to use `na.rm = TRUE` to remove NA values.

```{r}
# calculate the mean departure delay for all the flights. Remove NA.
summarize(flights, delay = mean(dep_delay, na.rm = TRUE))
```

You  use `nrow()`, `filter()`, and `is.na()` check the number of `NA` values for a column.

```{r}
# total rows in flights
nrow(flights)
# [1] 336776

# rows where dep_delay is NA
nrow(filter(flights, is.na(dep_delay)))
# [1] 8255

# rows where dep_delay has a value
nrow(filter(flights, !is.na(dep_delay)))
# [1] 328521
```


Use `summarize()` and `group_by()` to calculate values for each group.

```{r}
# group flights by year, month and day to get daily flights
by_day <- group_by(flights, year, month, day)

# calculate mean departure delay for each day
summarize(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

### Combining multiple operations with the pipe

Use pipe `%>%` to perform multiple operations on a data set. Do step 1, and then do step 2...

pipe makes code more readable.

```{r}
# groyup flights by day, and then calculate mean departure delay for each day
flights %>%
  group_by(year, month, day) %>%
  summarize(delay = mean(dep_delay, na.rm = TRUE))
```

### Counts

Whenever you do any aggregation, you should include a count of values `n()`, or a count of non-missing values `sum(!is.na(x))` to check number of items per group.

```{r}

#  get  flights that were not cancelled
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))

# group flights by tail number, and then calculate mean arrival delay and number of not canceled flights for each tail number
delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

delays
```

combine dplyr with ggplot to rearrange and plot data.

dplyr and rest of tidyverse uses `%\>% `; ggplot uses `+`

```{r}

# pick the delayed flights with more than 25 flights, and then plot the flights.
delays %>%
  filter(n > 25) %>%
  ggplot(mapping = aes(x = n, y = delay)) +
    geom_point(alpha = 1/10)

```

### Useful summary functions

Many functions can be used with `summarize()`

Measures of location: `median(x)` value where 50% of x is above it, and 50% is below it.

Measures of spread: `sd(x)` standard deviation is the standard measure of spread. interquartile range `IQR(x)` and median absolute deviation `mad(x)` are good if there are outliers.

Measures of rank: `min(x)`, `quantile(x, 0.25)`, `max(x)`

Measures of position: `first(x)`, `nth(x, 2)`, `last(x)`.

Counts: `n()` returns the size of the current group. `sum(!is.na(x))` returns the number of non-missing values. `n_distinct(x)` returns the number of unique values.

`count()` is a shortcut for `group_by()` and `summarize()` to return count by group.

```{r}
# count the number of flights per destination
not_cancelled %>%
  group_by(dest) %>%
  summarize(n = n())
```

```{r}
# count the number of flights per destination
not_cancelled %>%
  count(dest)
```

use `sort=TRUE` with `count()` to sort the counts

```{r}
# count the number of flights per destination, and sort the results
not_cancelled %>%
  count(dest, sort=TRUE)
```

optionally provide a weight variable to get a sum of that variable.

```{r}
# get the total number of miles planes flew for each destination
not_cancelled %>%
  count(dest, wt = distance)
```

When used with numeric functions, `TRUE` is converted to 1 and `FALSE` is converted to 0.

```{r, eval=FALSE}
sum(TRUE)
# [1] 1
sum(5 > 0)
# [1] 1


sum(FALSE)
# [1] 0
sum(5 < 0)
# [1] 0
```

Counts and proportions of logical values: `sum(x)` gives the number of TRUEs, and `mean(x)` gives the proportion of TRUEs.

```{r}
#  number of flight per day delayed by more than an 60 minutes
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(hour_count = sum(arr_delay > 60))
```

```{r}
#  proportion of flights per day delayed by more than an 60 minutes
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(hour_prop = mean(arr_delay > 60))
```

### Grouping by multiple variables

progressively rolling up summaries for sums and counts

```{r}
# group flights by day
daily <- group_by(flights, year, month, day)
```

```{r}
# number flights per day
(per_day <- summarize(daily, flights = n()))
```

```{r}
 # number flights per month
(per_month <- summarize(per_day, flights = sum(flights)))
```

```{r}
# number of flights per year
(per_year  <- summarize(per_month, flights = sum(flights)))
```

### Ungrouping

`ungroup()` removes grouping.

```{r}
# ungroup the daily flights to count the total number of flights
daily %>%
  ungroup() %>%
  summarize(flights = n())
```

If during analysis the data doesn't look right, you can try to use `ungroup()` to check if the data was previously grouped.

## Grouped mutates (and filters)

use `group()` with `mutate()` and `filter()`


```{r}
# get the flights with the top 10  largest arrival delays per day
top_delay <- flights_7_columns %>%
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) <= 10)
top_delay
```

```{r}
# add rank column, and sort the rows by day and rank to check if top_delay
# is returning what we expect
top_delay %>%
  mutate(rank = rank(desc(arr_delay)) ) %>%
  arrange(year, month, day, rank)
```


