# Workflow: code style

*"Good coding style is like correct punctuation: you can   manage without it, butitsuremakesthingseasiertoread." ~ Tidyverse Style Guide*

**Learning objectives:**
  
- Introduce the most important point's of the **tidyverse style** guide.

* Good coding style is synonymous to correct punctuation in writing.
    + In this chapter well learn how to:
        - Select good variable names.
        - Appropriately utilize spaces ` ` between mathematical operators.
        - Apply the `|>` operator to make our code more readable.
    + **Lorenz Walthert's** styler package is a good place to start from
        - Install it with install.packages("styler"), an easy way to use it, is via RStudio’s command palette.
        - Open the palette by pressing **Cmd/Ctrl + Shift + P**, then type **“styler”** to see all the shortcuts provided by styler.


```{r eval=FALSE}
library(tidyverse)
library(nycflights13)
```



## Names

Remember that variable names (those created by `<-` and those created by `mutate()`) should use only:

1. lowercase letters
2. numbers 
3. underscore

Use underscore to separate words within a name (also known as `snake_case`):

```{r eval=FALSE}
# Strive for:
short_flights <- flights |> filter(air_time < 60)

# Avoid:
SHORTFLIGHTS <- flights |> filter(air_time < 60)
```

Prefer long, descriptive names that are easy to understand, rather than concise names that are fast to type.\

If you have a bunch of names for related things, do your best to be consistent.\


## Spaces

*Put spaces on either side of mathematical operators apart from `^` (i.e., `+`, `-`, `==`, `<`, …), and around the assignment operator (`<-`).\

```{r eval=FALSE}
# Strive for
z <- (a + b)^2 / d

# Avoid
z<-( a + b ) ^ 2/d
```

Don’t put spaces inside or outside parentheses for regular function calls.\

**Always put a space after a comma, just like in regular English.**\

```{r eval=FALSE}
# Strive for
mean(x, na.rm = TRUE)

# Avoid
mean (x ,na.rm=TRUE)
```

It’s OK to add extra spaces if it improves alignment.\

For example, if you’re creating multiple variables in `mutate()`, you might want to add spaces so that all the `=` line up.\

This makes it easier to skim the code.\

```{r eval=FALSE}
flights |> 
  mutate(
    speed      = air_time / distance,
    dep_hour   = dep_time %/% 100,
    dep_minute = dep_time %%  100
  )
```


## Pipes

`|>` should always have a space before it and should typically be the last thing on a line.

This makes it easier to:

1. add new steps
2. rearrange existing steps
3. modify elements within a step
4. to get a 50,000 ft view by skimming the verbs on the left-hand side

```{r eval=FALSE}
# Strive for 
flights |>  
  filter(!is.na(arr_delay), !is.na(tailnum)) |> 
  count(dest)

# Avoid
flights|>filter(!is.na(arr_delay), !is.na(tailnum))|>count(dest)
```

If the function you’re piping into has named arguments (like `mutate()` or `summarize()`), put each argument on a new line.

If the function doesn’t have named arguments (like `select()` or `filter()`) keep everything on one line unless it doesn’t fit, in which case you should put each argument on its own line.

```{r eval=FALSE}
# Strive for
flights |>  
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights |>
  group_by(
    tailnum
  ) |> 
  summarize(delay = mean(arr_delay, na.rm = TRUE), n = n())
```

After the first step of the pipeline, indent each line by two spaces.\

```{r eval=FALSE}
# Strive for 
flights |>  
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights|>
  group_by(tailnum) |> 
  summarize(
             delay = mean(arr_delay, na.rm = TRUE), 
             n = n()
           )

flights|>
  group_by(tailnum) |> 
  summarize(
  delay = mean(arr_delay, na.rm = TRUE), 
  n = n()
  )
```

It’s OK to shirk some of these rules if your pipeline fits easily on one line.\ 

```{r eval = FALSE}
# This fits compactly on one line
df |> mutate(y = x + 1)

# While this takes up 4x as many lines, it's easily extended to 
# more variables and more steps in the future
df |> 
  mutate(
    y = x + 1
  )
```

Try to Limit your pipes around 10-15 lines.\

Try to break them up into smaller sub-tasks, giving each task an informative name.\

## ggplot2

The same basic rules that apply to the pipe also apply to `ggplot2`; just treat `+` the same way as `|>`.

```{r eval=FALSE}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE)
  ) |> 
  ggplot(aes(month, delay)) +
  geom_point() + 
  geom_line()
```

Again, if you can fit all of the arguments to a function on to a single line, put each argument on its own line:\

```{r eval=FALSE}
flights |> 
  group_by(dest) |> 
  summarize(
    distance = mean(distance),
    speed = mean(air_time / distance, na.rm = TRUE)
  ) |> 
  ggplot(aes(distance, speed)) +
  geom_smooth(
    method = "loess",
    span = 0.5,
    se = FALSE, 
    color = "white", 
    size = 4
  ) +
  geom_point()
```

## Selectioning comments

As your scripts get longer, you can use sectioning comments to break up your file into manageable pieces:\

```{r load data comment}
 # Load data --------------------------------------

 # Plot data --------------------------------------
```

A keyboard shortcut to create these headers is (**Cmd/Ctrl + Shift + R**)\

## Exercises

1. Restyle the following pipelines following the guidelines above.

```{r exercises, eval=FALSE}
flights|>filter(dest=="IAH")|>group_by(year,month,day)|>summarize(n=n(),
delay=mean(arr_delay,na.rm=TRUE))|>filter(n>10)

flights|>filter(carrier=="UA",dest%in%c("IAH","HOU"),sched_dep_time>
0900,sched_arr_time<2000)|>group_by(flight)|>summarize(delay=mean(
arr_delay,na.rm=TRUE),cancelled=sum(is.na(arr_delay)),n=n())|>filter(n>10)
```

Solution

```{r exercises solution, eval=FALSE}
flights |> # space before/ after pipe and the last thing on a line
  filter(dest == "IAH") |> # indent twice, space between operators
  group_by(year, month, day) |> # space after comma
  summarize( # nam
    n     = n(), # add spaces so that all the = line up
    delay = mean(arr_delay, na.rm = TRUE)
    ) |> 
  filter(n > 10)


flights |> 
  filter(
         carrier        == "UA", 
         dest           %in% c("IAH","HOU"),
         sched_dep_time > 900,
         sched_arr_time < 2000
         ) |> 
  group_by(flight) |> 
  summarize(
            delay = mean(arr_delay, na.rm=TRUE),
            cancelled = sum(is.na(arr_delay)),
            n = n()
            ) |> 
  filter(n > 10)
```


## Meeting Videos

### Cohort 5 

`r knitr::include_url("https://www.youtube.com/embed/JW1DilkJ3-0")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:12:57	Jon Harmon (jonthegeek):	https://en.wikipedia.org/wiki/The_Treachery_of_Images
00:16:47	Jon Harmon (jonthegeek):	https://cran.r-project.org/package=magrittr
00:20:53	Jon Harmon (jonthegeek):	https://smile.amazon.com/Mathematical-Notation-Guide-Engineers-Scientists/dp/1466230525/ref=sr_1_2?keywords=Mathematical+Notation&qid=1641665745&sr=8-2
00:22:16	Jon Harmon (jonthegeek):	|>
00:23:04	Ryan Metcalf:	And, is this the “Walrus” operator? `:=`
00:23:19	Jon Harmon (jonthegeek):	%>%
00:23:28	Jon Harmon (jonthegeek):	|>
00:24:14	Jon Harmon (jonthegeek):	a |> f()
00:24:23	Jon Harmon (jonthegeek):	f(a)
00:41:01	Becki R. (she/her):	Ryan, can you put a link to the video in the chat?
00:42:27	Jon Harmon (jonthegeek):	a %>% myfun(a = 10, b = .)
00:42:59	Jon Harmon (jonthegeek):	a |> myfun(a = 10, b = .)
00:43:00	Ryan Metcalf:	Jenny Bryant Youtube Video: https://youtu.be/4MfUCX_KpdE
00:43:07	Becki R. (she/her):	Thanks
00:44:57	Jon Harmon (jonthegeek):	a %>% myfun(arg1 = 10, arg2 = .)
00:45:48	Jon Harmon (jonthegeek):	a %>% myfun(arg2 = .)
00:47:40	Federica Gazzelloni:	https://www.r-bloggers.com/2021/05/the-new-r-pipe/
00:50:00	Jon Harmon (jonthegeek):	https://twitter.com/hadleywickham/status/1359852563726819332
00:51:58	Federica Gazzelloni:	https://community.rstudio.com/t/why-is-r-core-creating-a-new-pipe-operator/89020
01:03:35	Federica Gazzelloni:	you can use: set.seed(123) with some numbers inside to recall the same random set
01:03:56	Federica Gazzelloni:	set.seed(123)
01:04:06	Federica Gazzelloni:	rnorm(100)…
01:08:46	Jon Harmon (jonthegeek):	a <- a %>%
a %<>%
01:10:41	Jon Harmon (jonthegeek):	mtcars <- head(mtcars)
01:10:50	Jon Harmon (jonthegeek):	mtcars <- mtcars %>% head()
01:10:57	Jon Harmon (jonthegeek):	mtcars %<>% head()
01:13:56	Becki R. (she/her):	Where is the signup list for presenting chapters?
```
</details>

### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/jwnQkHtUiOY")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:09:32	Marielena Soilemezidi:	=
00:26:49	Daniel Adereti:	Yes, I agree. It appears the dot allows to insert any argument
00:52:38	Daniel Adereti:	Thanks Marielena, also share Shannon's thoughts, pipes had always been taught like a must use operator, glad to see it is meant to be selective
00:55:20	Daniel Adereti:	works for me too, it is quite a large chapter
00:55:39	Daniel Adereti:	Sorry, my mic seems problematic, hence my typing
```
</details>

### Cohort 7

`r knitr::include_url("https://www.youtube.com/embed/wlYh2KewQcI")`

<details>
  <summary> Meeting chat log </summary>
```
00:21:34	Oluwafemi Oyedele:	https://towardsdatascience.com/understanding-the-native-r-pipe-98dea6d8b61b
00:22:15	Christine (she/her):	Thanks, Oluwafemi!
00:22:51	Oluwafemi Oyedele:	https://stackoverflow.com/questions/67633022/what-are-the-differences-between-rs-new-native-pipe-and-the-magrittr-pipe
00:37:22	Oluwafemi Oyedele:	https://github.com/daranzolin/ViewPipeSteps
00:49:42	timnewby:	https://towardsdatascience.com/an-introduction-to-the-pipe-in-r-823090760d64 Talks a bit about when pipes work well and when they might not
00:50:13	Oluwafemi Oyedele:	http://127.0.0.1:22120/session/Rvig.11e05a207606.html
01:03:47	Dolleen Osundwa:	thank you Kumar
```
</details>

`r knitr::include_url("https://www.youtube.com/embed/LHYUEAXaYoM")`


### Cohort 8

`r knitr::include_url("https://www.youtube.com/embed/uEYKR3xs4Y0")`

`r knitr::include_url("https://www.youtube.com/embed/MvzrbyCuFLw")`
