# Workflow: code style

*Note: This was previously 2 separate chapters. Update the slides for the present single-chapter setup, and delete this note.*

**Learning objectives:**

- Use the `%>%` operator to make code more readable.
  - `lhs` versus `rhs`
- Recognize when *not* to use the `%>%` operator.
- Review other pipe operators that may be helpful.

## Introduction

- The pipe operator comes from the `{magrittr}` package by [Stefan Milton Bache](https://stefanbache.dk/)
- If you use ***ANY*** tidyverse libraries, `{magrittr}` is already included (how to validate this statement)
- For learning humans... read the `%>%` operator as semantically: *and then*

```{r, 05-Loading magrittr, eval=FALSE}
library(magrittr)
```


## Lets see it in Code!!!

The following example will highlight a code snippet **without** piping (Object Assignment and Nested) and **with** piping.

```{r, 05-Example without Pipe, eval = FALSE}
a <- dplyr::filter(mtcars, carb > 1)
b <- dplyr::group_by(a, cyl)
c <- dplyr::summarise(b, Avg_mpg = mean(mpg))
d <- dplyr::arrange(c, desc(Avg_mpg))
print(d)
```
Above, objects `a-d` are created as "placeholders". Objects that are nothing more then memory space to hold the output of a function. The difference is, each of these can be called upon in future script.

> One of my favorite talks: [Jenny Bryant at Plotcon, 2016](https://youtu.be/4MfUCX_KpdE). Specific to the topic, start at min 15:00. The discussion is related to Dataframes and the key takeaway is wrangle your data once.

But couldn't I just nest my function?
```{r, 05-Nested Call, eval = FALSE}
dplyr::arrange(
   dplyr::summarize(
       dplyr::group_by(
           dplyr::filter(mtcars, carb > 1),
           cyl
       ),
       Avg_mpg = mean(mpg)
   ),
   desc(Avg_mpg)
 )
```
Nesting a long function is acceptable and in some languages preferred. However, given the structure of RStudio and its plethora of packages, there is no need.

Using the Pipe
```{r, 05-Example with Pipe, eval=FALSE}
mtcars %>%
        dplyr::filter(carb > 1) %>%
        dplyr::group_by(cyl) %>%
        dplyr::summarise(Avg_mpg = mean(mpg)) %>%
        dplyr::arrange(desc(Avg_mpg))
```
The third example not only reduces complexity... you are building a pipeline to process the `mtcars` dataset. There is no difference in computation from the first example... only, now everything is in memory.

>NOTE: There is a section in here about `pryr` that I wasn't sure the intent other then a discussion on memory use. R is intellegent and will share colums across dataframes. The material was excluded from the discussion.

## Overwritting Data... easy, right?

One method of wrangling data is to just overwrite and reassign a calculation to its original object. But, there is a problem...

1. Overwriting an object prevents clean debugging. You'll have to rerun your code over and over to see what is changing.
2. The repetition of the object being transformed implies... "what does my data look like now?"

If you want to make yourself crazy... continually read in a new copy of your data just to walk through the process over again. This is *not* conducive to efficiency.

## The pipe makes lexical transformations

- `magrittr` reassembles the code in the pipe to a form that works by overwriting an intermediate object, see [link](https://adv-r.hadley.nz/functions.html#lexical-scoping)
- This means that the pipe won’t work for two classes of functions:
 
 1. Functions that use the current environment.
 2. Functions that use lazy evaluation.
 
## When not to use the pipe

- Your pipes are longer than (say) ten steps.
- You have multiple inputs or outputs.
- You are starting to think about a directed graph with a complex dependency structure.

## Other tools from `magrittr`

Before we begin, let's discuss `lhs` and `rhs`

- `lhs` - Left Hand Side
- `rhs` - Right Hand Side

Alternative Pipes

- **Tee** pipe: `%T>%`
  - When working with more complex pipes, it's sometimes useful to call a function for its side-effects
    - Returns the left-hand side instead of the right-hand side
  
```{r, 05-Tee Pipe, include=TRUE}
library(magrittr)
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()

# Using the Tee pipe allows for seeing the data prior to plotting
rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
```

- **Exposition** pipe: `%$%`
  - It *explodes* out the variables in a data frame so that you can refer to them explicitly
 
```{r, 05-Exposition Pipe, include = TRUE}
# Output the correlation of `disp` to `mpg`
mtcars %$%
  cor(disp, mpg)
```
 
- **Assignment** pipe: `%<>%`
  - Allows you to replace code
 
```{r, 05-Without Assignment, include = TRUE}
mtcars <- mtcars %>% 
  transform(cyl = cyl * 2)
```

Using the Assignment Operator `%<>%`

```{r, 05-With Assignment, include = TRUE}
mtcars %<>% transform(cyl = cyl * 2)
```

> *In my opinion, a little bit of duplication (i.e. repeating the name of the object twice) is fine in return for making assignment more explicit.* - Hadley Wickham

## Meeting Videos

### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/JW1DilkJ3-0")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:12:57	Jon Harmon (jonthegeek):	https://en.wikipedia.org/wiki/The_Treachery_of_Images
00:16:47	Jon Harmon (jonthegeek):	https://cran.r-project.org/package=magrittr
00:20:53	Jon Harmon (jonthegeek):	https://smile.amazon.com/Mathematical-Notation-Guide-Engineers-Scientists/dp/1466230525/ref=sr_1_2?keywords=Mathematical+Notation&qid=1641665745&sr=8-2
00:22:16	Jon Harmon (jonthegeek):	|>
00:23:04	Ryan Metcalf:	And, is this the “Walrus” operator? `:=`
00:23:19	Jon Harmon (jonthegeek):	%>%
00:23:28	Jon Harmon (jonthegeek):	|>
00:24:14	Jon Harmon (jonthegeek):	a |> f()
00:24:23	Jon Harmon (jonthegeek):	f(a)
00:41:01	Becki R. (she/her):	Ryan, can you put a link to the video in the chat?
00:42:27	Jon Harmon (jonthegeek):	a %>% myfun(a = 10, b = .)
00:42:59	Jon Harmon (jonthegeek):	a |> myfun(a = 10, b = .)
00:43:00	Ryan Metcalf:	Jenny Bryant Youtube Video: https://youtu.be/4MfUCX_KpdE
00:43:07	Becki R. (she/her):	Thanks
00:44:57	Jon Harmon (jonthegeek):	a %>% myfun(arg1 = 10, arg2 = .)
00:45:48	Jon Harmon (jonthegeek):	a %>% myfun(arg2 = .)
00:47:40	Federica Gazzelloni:	https://www.r-bloggers.com/2021/05/the-new-r-pipe/
00:50:00	Jon Harmon (jonthegeek):	https://twitter.com/hadleywickham/status/1359852563726819332
00:51:58	Federica Gazzelloni:	https://community.rstudio.com/t/why-is-r-core-creating-a-new-pipe-operator/89020
01:03:35	Federica Gazzelloni:	you can use: set.seed(123) with some numbers inside to recall the same random set
01:03:56	Federica Gazzelloni:	set.seed(123)
01:04:06	Federica Gazzelloni:	rnorm(100)…
01:08:46	Jon Harmon (jonthegeek):	a <- a %>%
a %<>%
01:10:41	Jon Harmon (jonthegeek):	mtcars <- head(mtcars)
01:10:50	Jon Harmon (jonthegeek):	mtcars <- mtcars %>% head()
01:10:57	Jon Harmon (jonthegeek):	mtcars %<>% head()
01:13:56	Becki R. (she/her):	Where is the signup list for presenting chapters?
```
</details>

### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/jwnQkHtUiOY")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:09:32	Marielena Soilemezidi:	=
00:26:49	Daniel Adereti:	Yes, I agree. It appears the dot allows to insert any argument
00:52:38	Daniel Adereti:	Thanks Marielena, also share Shannon's thoughts, pipes had always been taught like a must use operator, glad to see it is meant to be selective
00:55:20	Daniel Adereti:	works for me too, it is quite a large chapter
00:55:39	Daniel Adereti:	Sorry, my mic seems problematic, hence my typing
```
</details>



## Previously Chapter 7 {-}

**Learning objectives:**


Introduce the most important point's of the **tidyverse style** guide.\

Good coding style is synonymous to correct punctuation in writing.\

Styling your code will feel a bit tedious to start with, but if you practice it, it will soon become second nature.\

**Lorenz Walthert's** styler package is a good place to start from\

Install it with install.packages("styler"), an easy way to use it, is via RStudio’s command palette.\

Open the palette by pressing **Cmd/Ctrl + Shift + P**, then type **“styler”** to see all the shortcuts provided by styler.\


```{r eval=FALSE}
library(tidyverse)
library(nycflights13)
```

## Names

Remember that variable names (those created by <- and those created by mutate()) should use only:\
  1. lowercase letters,\
  2. numbers, and underscore\
Use underscore  to separate words within a name\

```{r eval=FALSE}
# Strive for:
short_flights <- flights |> filter(air_time < 60)

# Avoid:
SHORTFLIGHTS <- flights |> filter(air_time < 60)
```

Prefer long, descriptive names that are easy to understand, rather than concise names that are fast to type.\

If you have a bunch of names for related things, do your best to be consistent.\


## Spaces

*Put spaces on either side of mathematical operators apart from ^ (i.e., +, -, ==, <, …), and around the assignment operator (<-).\

```{r eval=FALSE}
# Strive for
z <- (a + b)^2 / d

# Avoid
z<-( a + b ) ^ 2/d
```

Don’t put spaces inside or outside parentheses for regular function calls.\

**Always put a space after a comma, just like in regular English.**\

```{r eval=FALSE}
# Strive for
mean(x, na.rm = TRUE)

# Avoid
mean (x ,na.rm=TRUE)
```

It’s OK to add extra spaces if it improves alignment.\

For example, if you’re creating multiple variables in mutate(), you might want to add spaces so that all the = line up.\

This makes it easier to skim the code.\

```{r eval=FALSE}
flights |> 
  mutate(
    speed      = air_time / distance,
    dep_hour   = dep_time %/% 100,
    dep_minute = dep_time %%  100
  )
```


## Pipes

|> should always have a space before it and should typically be the last thing on a line.\

This makes it easier to:\
  1.add new steps,\
  2.rearrange existing steps,\
  3.modify elements within a step,\
  4.and to get a 50,000 ft view by skimming the verbs on the      left-hand side.\

```{r eval=FALSE}
# Strive for 
flights |>  
  filter(!is.na(arr_delay), !is.na(tailnum)) |> 
  count(dest)

# Avoid
flights|>filter(!is.na(arr_delay), !is.na(tailnum))|>count(dest)
```

If the function you’re piping into has named arguments (like mutate() or summarize()), put each argument on a new line.\

If the function doesn’t have named arguments (like select() or filter()) keep everything on one line unless it doesn’t fit, in which case you should put each argument on its own line.\

```{r eval=FALSE}
# Strive for
flights |>  
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights |>
  group_by(
    tailnum
  ) |> 
  summarize(delay = mean(arr_delay, na.rm = TRUE), n = n())
```

After the first step of the pipeline, indent each line by two spaces.\

```{r eval=FALSE}
# Strive for 
flights |>  
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights|>
  group_by(tailnum) |> 
  summarize(
             delay = mean(arr_delay, na.rm = TRUE), 
             n = n()
           )

flights|>
  group_by(tailnum) |> 
  summarize(
  delay = mean(arr_delay, na.rm = TRUE), 
  n = n()
  )
```

It’s OK to shirk some of these rules if your pipeline fits easily on one line.\ 

```{r eval = FALSE}
# This fits compactly on one line
df |> mutate(y = x + 1)

# While this takes up 4x as many lines, it's easily extended to 
# more variables and more steps in the future
df |> 
  mutate(
    y = x + 1
  )
```

Try to Limit your pipes around 10-15 lines.\

Try to break them up into smaller sub-tasks, giving each task an informative name.\

## ggplot2

**The same basic rules that apply to the pipe also apply to ggplot2; just treat + the same way as |>.**\

```{r eval=FALSE}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE)
  ) |> 
  ggplot(aes(month, delay)) +
  geom_point() + 
  geom_line()
```

Again, if you can fit all of the arguments to a function on to a single line, put each argument on its own line:\

```{r eval=FALSE}
flights |> 
  group_by(dest) |> 
  summarize(
    distance = mean(distance),
    speed = mean(air_time / distance, na.rm = TRUE)
  ) |> 
  ggplot(aes(distance, speed)) +
  geom_smooth(
    method = "loess",
    span = 0.5,
    se = FALSE, 
    color = "white", 
    size = 4
  ) +
  geom_point()
```

## Selectioning comments

As your scripts get longer, you can use sectioning comments to break up your file into manageable pieces:\

 # Load data --------------------------------------

 # Plot data --------------------------------------

A keyboard shortcut to create these headers is (Cmd/Ctrl + Shift + R)\



## Meeting Videos

### Cohort 7

`r knitr::include_url("https://www.youtube.com/embed/wlYh2KewQcI")`

<details>
  <summary> Meeting chat log </summary>
```
00:21:34	Oluwafemi Oyedele:	https://towardsdatascience.com/understanding-the-native-r-pipe-98dea6d8b61b
00:22:15	Christine (she/her):	Thanks, Oluwafemi!
00:22:51	Oluwafemi Oyedele:	https://stackoverflow.com/questions/67633022/what-are-the-differences-between-rs-new-native-pipe-and-the-magrittr-pipe
00:37:22	Oluwafemi Oyedele:	https://github.com/daranzolin/ViewPipeSteps
00:49:42	timnewby:	https://towardsdatascience.com/an-introduction-to-the-pipe-in-r-823090760d64 Talks a bit about when pipes work well and when they might not
00:50:13	Oluwafemi Oyedele:	http://127.0.0.1:22120/session/Rvig.11e05a207606.html
01:03:47	Dolleen Osundwa:	thank you Kumar
```
</details>

`r knitr::include_url("https://www.youtube.com/embed/LHYUEAXaYoM")`


### Cohort 8

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary> Meeting chat log </summary>
```
LOG
```
</details>
