# Exploratory Data Analysis

## Learning objectives

```{r 07-setup, include = FALSE}
library(tidyverse)
```

- Recognize the **two types of questions** that will always be useful for making discoveries within your data: "What type of variation occurs within my variables?" and "What type of covariation occurs between my variables?"
- Explore the **variation** within the variables of your observations.
- Deal with outliers and **missing values** in your data.
- Explore the **covariation** between the variables of your observations.
- Recognize how **models** can be used to explore **patterns** in your data.

## Overall Vocabulary

- **variable:** a quantity, quality, or property that you can measure.
- **value:** the state of a variable when you measure it. Can change.
- **observation:** a set of measurements made under similar conditions. One value per variable.
- **tabular data:** observations of variables.
- **tidy data:** 1 observation per row, 1 variable per column, 1 value per cell. Definition of "tidy" for a dataset can depend on what you're trying to answer.

## Variation

- **variation:** the tendency of values of a variable to change between measurements.
- **categorical variable:** can only take certain values. Visualize variation with bar chart.

```{r 07-variation-categorical}
ggplot(data = diamonds) +
  aes(x = cut) +
  geom_bar()
```

- **continuous variables:** can take on infinite set of ordered values. Visualize variation with histogram.

```{r 07-variation-continuous}
ggplot(data = diamonds) +
  aes(x = carat) +
  geom_histogram(binwidth = 0.5)
```

- `geom_freqpoly` is `geom_histogram` alternative that doesn't show bars.
- Reminder: the `%>%` pipe = "and then".
- `{ggplot2}` uses `+` to add layers, read it as "with" or "and".

```{r 07-variation-freqpoly}
smaller <- diamonds %>% 
  filter(carat < 3) 
ggplot(smaller) +
  aes(x = carat, colour = cut) +
  geom_freqpoly(binwidth = 0.1)
```

- Use the visualizations to develop questions!
  - Which values are the most common? Why?
  - Which values are rare? Why? Does that match your expectations?
  - Can you see any unusual patterns? What might explain them?

```{r 07-var-questions1}
ggplot(smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

- Subgroups create more questions:
  - How are the observations within each cluster similar to each other?
  - How are the observations in separate clusters different from each other?
  - How can you explain or describe the clusters?
  - Why might the appearance of clusters be misleading?

- Use `coord_cartesian` to zoom in to see unusual values.
- Can be ok to drop weird values, especially if you can explain where they came from.
- Always disclose that you did that, though.

## Missing values

2 options to deal with weird values:

- Drop the entire row. <-- probably don't do this
- Replace bad data with NA.

```{r 07-replace-weird}
diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))
```

- `{ggplot2}` will give a warning when values are missing, can suppress with `na.rm = TRUE`.

## Covariation

- **covariation:** tendency of values of different variables to vary *together* in a related way.
- Visualizing covariance depends on types of variables in the pair:

### categorical + continuous

- `x = categorical`
- `y = continuous`.
- `geom_boxplot`
- Lots of options exist to do this better. See [Cedric Scherer's tutorial](https://www.cedricscherer.com/2021/06/06/visualizing-distributions-with-raincloud-plots-with-ggplot2/)!

### categorical + categorical

- `geom_count`
- `dplyr::count` then `geom_tile`

### continuous + continuous

- `geom_point`
- `geom_bin2d`
- `geom_hex`


## Finding Patterns

**Ask yourself:**

- Could this pattern be due to coincidence (i.e. random chance)?
- How can you describe the relationship implied by the pattern?
- How strong is the relationship implied by the pattern?
- What other variables might affect the relationship?
- Does the relationship change if you look at individual subgroups of the data?

## Simplified ggplot2

```{r 07-ggplot-simplified}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_freqpoly(binwidth = 0.25)

ggplot(faithful, aes(eruptions)) + 
  geom_freqpoly(binwidth = 0.25)

# Or Jon's crazy way
ggplot(faithful) +
  aes(eruptions) +
  geom_freqpoly(binwidth = 0.25)
```

## Learning More

- [r4ds.io/join](r4ds.io/join) for more book clubs!