# Workflow: code style

**Learning objectives:**


Introduce the most important point's of the **tidyverse style** guide.\

Good coding style is synonymous to correct punctuation in writing.\

Styling your code will feel a bit tedious to start with, but if you practice it, it will soon become second nature.\

**Lorenz Walthert's** styler package is a good place to start from\

Install it with install.packages("styler"), an easy way to use it, is via RStudio’s command palette.\

Open the palette by pressing **Cmd/Ctrl + Shift + P**, then type **“styler”** to see all the shortcuts provided by styler.\


```{r eval=FALSE}
library(tidyverse)
library(nycflights13)
```

## Names

Remember that variable names (those created by <- and those created by mutate()) should use only:\
  1. lowercase letters,\
  2. numbers, and underscore\
Use underscore  to separate words within a name\

```{r eval=FALSE}
# Strive for:
short_flights <- flights |> filter(air_time < 60)

# Avoid:
SHORTFLIGHTS <- flights |> filter(air_time < 60)
```

Prefer long, descriptive names that are easy to understand, rather than concise names that are fast to type.\

If you have a bunch of names for related things, do your best to be consistent.\


## Spaces

*Put spaces on either side of mathematical operators apart from ^ (i.e., +, -, ==, <, …), and around the assignment operator (<-).\

```{r eval=FALSE}
# Strive for
z <- (a + b)^2 / d

# Avoid
z<-( a + b ) ^ 2/d
```

Don’t put spaces inside or outside parentheses for regular function calls.\

**Always put a space after a comma, just like in regular English.**\

```{r eval=FALSE}
# Strive for
mean(x, na.rm = TRUE)

# Avoid
mean (x ,na.rm=TRUE)
```

It’s OK to add extra spaces if it improves alignment.\

For example, if you’re creating multiple variables in mutate(), you might want to add spaces so that all the = line up.\

This makes it easier to skim the code.\

```{r eval=FALSE}
flights |> 
  mutate(
    speed      = air_time / distance,
    dep_hour   = dep_time %/% 100,
    dep_minute = dep_time %%  100
  )
```


## Pipes

|> should always have a space before it and should typically be the last thing on a line.\

This makes it easier to:\
  1.add new steps,\
  2.rearrange existing steps,\
  3.modify elements within a step,\
  4.and to get a 50,000 ft view by skimming the verbs on the      left-hand side.\

```{r eval=FALSE}
# Strive for 
flights |>  
  filter(!is.na(arr_delay), !is.na(tailnum)) |> 
  count(dest)

# Avoid
flights|>filter(!is.na(arr_delay), !is.na(tailnum))|>count(dest)
```

If the function you’re piping into has named arguments (like mutate() or summarize()), put each argument on a new line.\

If the function doesn’t have named arguments (like select() or filter()) keep everything on one line unless it doesn’t fit, in which case you should put each argument on its own line.\

```{r eval=FALSE}
# Strive for
flights |>  
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights |>
  group_by(
    tailnum
  ) |> 
  summarize(delay = mean(arr_delay, na.rm = TRUE), n = n())
```

After the first step of the pipeline, indent each line by two spaces.\

```{r eval=FALSE}
# Strive for 
flights |>  
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights|>
  group_by(tailnum) |> 
  summarize(
             delay = mean(arr_delay, na.rm = TRUE), 
             n = n()
           )

flights|>
  group_by(tailnum) |> 
  summarize(
  delay = mean(arr_delay, na.rm = TRUE), 
  n = n()
  )
```

It’s OK to shirk some of these rules if your pipeline fits easily on one line.\ 

```{r eval = FALSE}
# This fits compactly on one line
df |> mutate(y = x + 1)

# While this takes up 4x as many lines, it's easily extended to 
# more variables and more steps in the future
df |> 
  mutate(
    y = x + 1
  )
```

Try to Limit your pipes around 10-15 lines.\

Try to break them up into smaller sub-tasks, giving each task an informative name.\

## ggplot2

**The same basic rules that apply to the pipe also apply to ggplot2; just treat + the same way as |>.**\

```{r eval=FALSE}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE)
  ) |> 
  ggplot(aes(month, delay)) +
  geom_point() + 
  geom_line()
```

Again, if you can fit all of the arguments to a function on to a single line, put each argument on its own line:\

```{r eval=FALSE}
flights |> 
  group_by(dest) |> 
  summarize(
    distance = mean(distance),
    speed = mean(air_time / distance, na.rm = TRUE)
  ) |> 
  ggplot(aes(distance, speed)) +
  geom_smooth(
    method = "loess",
    span = 0.5,
    se = FALSE, 
    color = "white", 
    size = 4
  ) +
  geom_point()
```

## Selectioning comments

As your scripts get longer, you can use sectioning comments to break up your file into manageable pieces:\

 # Load data --------------------------------------

 # Plot data --------------------------------------

A keyboard shortcut to create these headers is (Cmd/Ctrl + Shift + R)\



## Meeting Videos

### Cohort 7

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary> Meeting chat log </summary>
```
LOG
```
</details>

### Cohort 8

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary> Meeting chat log </summary>
```
LOG
```
</details>
