# Exploratory Data Analysis

## Learning objectives

```{r}
#| label: 12-setup
#| include: false

library(tidyverse)
```

- Recognize the **two types of questions** that will always be useful for making discoveries within your data: "What type of variation occurs within my variables?" and "What type of covariation occurs between my variables?"
- Explore the **variation** within the variables of your observations.
- Deal with outliers and **missing values** in your data.
- Explore the **covariation** between the variables of your observations.
- Recognize how **models** can be used to explore **patterns** in your data.

## Overall Vocabulary

- **variable:** a quantity, quality, or property that you can measure.
- **value:** the state of a variable when you measure it. Can change.
- **observation:** a set of measurements made under similar conditions. One value per variable.
- **tabular data:** observations of variables.
- **tidy data:** 1 observation per row, 1 variable per column, 1 value per cell. Definition of "tidy" for a dataset can depend on what you're trying to answer.

## Variation

- **variation:** the tendency of values of a variable to change between measurements.
- **categorical variable:** can only take certain values. Visualize variation with bar chart.

```{r}
#| label: 12-variation-categorical

ggplot(data = diamonds) +
  aes(x = cut) +
  geom_bar()
```

- **continuous variables:** can take on infinite set of ordered values. Visualize variation with histogram.

```{r}
#| label: 12-variation-continuous

ggplot(data = diamonds) +
  aes(x = carat) +
  geom_histogram(binwidth = 0.5)
```

- `geom_freqpoly` is `geom_histogram` alternative that doesn't show bars.
- Reminder: the `%>%` pipe = "and then".
- `{ggplot2}` uses `+` to add layers, read it as "with" or "and".

```{r}
#| label: 12-variation-freqpoly

smaller <- diamonds %>% 
  filter(carat < 3) 
ggplot(smaller) +
  aes(x = carat, colour = cut) +
  geom_freqpoly(binwidth = 0.1)
```

- Use the visualizations to develop questions!
  - Which values are the most common? Why?
  - Which values are rare? Why? Does that match your expectations?
  - Can you see any unusual patterns? What might explain them?

```{r}
#| label: 12-var-questions1

ggplot(smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

- Subgroups create more questions:
  - How are the observations within each cluster similar to each other?
  - How are the observations in separate clusters different from each other?
  - How can you explain or describe the clusters?
  - Why might the appearance of clusters be misleading?

- Use `coord_cartesian` to zoom in to see unusual values.
- Can be ok to drop weird values, especially if you can explain where they came from.
- Always disclose that you did that, though.

## Missing values

2 options to deal with weird values:

- Drop the entire row. <-- probably don't do this
- Replace bad data with NA.

```{r}
#| label: 12-replace-weird

diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))
```

- `{ggplot2}` will give a warning when values are missing, can suppress with `na.rm = TRUE`.

## Covariation

- **covariation:** tendency of values of different variables to vary *together* in a related way.
- Visualizing covariance depends on types of variables in the pair:

### categorical + continuous

- `x = categorical`
- `y = continuous`.
- `geom_boxplot`
- Lots of options exist to do this better. See [Cedric Scherer's tutorial](https://www.cedricscherer.com/2021/06/06/visualizing-distributions-with-raincloud-plots-with-ggplot2/)!

### categorical + categorical

- `geom_count`
- `dplyr::count` then `geom_tile`

### continuous + continuous

- `geom_point`
- `geom_bin2d`
- `geom_hex`


## Finding Patterns

**Ask yourself:**

- Could this pattern be due to coincidence (i.e. random chance)?
- How can you describe the relationship implied by the pattern?
- How strong is the relationship implied by the pattern?
- What other variables might affect the relationship?
- Does the relationship change if you look at individual subgroups of the data?

## Simplified ggplot2

```{r}
#| label: 12-ggplot-simplified

ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_freqpoly(binwidth = 0.25)

ggplot(faithful, aes(eruptions)) + 
  geom_freqpoly(binwidth = 0.25)

# Or Jon's crazy way
ggplot(faithful) +
  aes(eruptions) +
  geom_freqpoly(binwidth = 0.25)
```

## Learning More

- [dslc.io/join](dslc.io/join) for more book clubs!
- [R Graph Gallery](https://www.r-graph-gallery.com/ggplot2-package.html)
- The [Graphs section](http://www.cookbook-r.com/Graphs/) of the R Cookbook

## Meeting Videos

### Cohort 5

{{< video https://www.youtube.com/embed/ujOn-4esnDo >}}

<details>
  <summary> Meeting chat log </summary>
  
```
00:13:43	Njoki Njuki Lucy:	Is it best to visualize the variation in a categorical variable with only two levels using a bar chart? If not, what's the chart to use if I may ask?
00:16:00	Ryan Metcalf:	Great question Njoki, Categorical, by definition is a set that a variable can have. Say, Male / Female / Other. This example indicates a variable can have three states. It depends on your data set.
00:16:51	Eileen:	bar or pie chart?
00:16:51	Ryan Metcalf:	There are other forms of presentation other than a bar chart. I.E “quantifying” each category.
00:18:37	Eileen:	box chart
00:18:46	Njoki Njuki Lucy:	thank you so much  everyone :)
00:24:31	lucus w:	This website is excellent in determining geom to use: www.data-to-viz.com
00:25:22	Njoki Njuki Lucy:	awesome, thanks
00:25:44	Eileen:	Box charts are great for showing outliers
00:26:31	Federica Gazzelloni:	other interesting resources:
00:26:34	Federica Gazzelloni:	https://www.r-graph-gallery.com/ggplot2-package.html
00:26:51	Federica Gazzelloni:	http://www.cookbook-r.com/Graphs/
00:34:19	Amitrajit:	what is the difference in putting aes()  inside geom_count() rather than main ggplot() call?
00:35:38	Ryan Metcalf:	Like maybe Supply vs Demand curves?
00:41:16	Federica Gazzelloni:	what about the factor() that we add to a variable when we apply a color?
00:42:33	Susie Neilson:	I do aes your way Jon!
00:43:07	Federica Gazzelloni:	and grouping inside the aes
00:49:27	Amitrajit:	thanks!
00:49:32	Federica Gazzelloni:	thanks
00:49:35	Njoki Njuki Lucy:	thank you, bye
00:49:45	Eileen:	Thank you!
```
</details>

### Cohort 6

{{< video https://www.youtube.com/embed/mYTD9DbM174 >}}

<details>
  <summary> Meeting chat log </summary>
  
```
00:06:21	Matthew Efoli:	good evening Daniel and Esmeralda
00:07:39	Matthew Efoli:	hello everyone
00:08:08	Daniel Adereti:	Hello Matthew!
00:08:44	Daniel Adereti:	I guess we can start? so we can finish the 2 chapters as Exploratory Data Analysis is quite long and involved
00:09:04	Freya Watkins:	Sounds good! Hi all :)
00:10:55	Freya Watkins:	yes can see
00:23:14	Daniel Adereti:	na > Not available
00:23:32	Maria Eleni Soilemezidi:	rm = remove
00:25:49	Esmeralda Cruz:	yes
00:26:29	Esmeralda Cruz:	to remove the outliers maybe?
00:29:20	Adeyemi Olusola:	No
00:29:22	Freya Watkins:	we can't see it no
00:29:27	Maria Eleni Soilemezidi:	no we can't see it!
00:29:38	Maria Eleni Soilemezidi:	thank you! Yes
00:32:57	Daniel Adereti:	Cedric's article is a nice one! Helpful to understand descriptive use case of different plot ideas
00:43:19	Daniel Adereti:	we can do the exercises
00:43:27	Esmeralda Cruz:	ok
00:43:28	Maria Eleni Soilemezidi:	yes, sure!
00:45:20	Adeyemi Olusola:	we can try reorder
00:45:28	Adeyemi Olusola:	from the previous example
00:51:44	Maria Eleni Soilemezidi:	that's a good idea
00:52:28	Daniel Adereti:	Thanks!
00:52:42	Daniel Adereti:	cut_in_color_graph <- diamonds %>% 
  group_by(color, cut) %>% 
  summarise(n = n()) %>% 
  mutate(proportion_cut_in_color = n/sum(n)) %>%
  ggplot(aes(x = color, y = cut))+
  geom_tile(aes(fill = proportion_cut_in_color))+
  labs(fill = "proportion\ncut in color")
00:53:32	Esmeralda Cruz:	😮
00:53:47	Adeyemi Olusola:	smiles
00:54:13	Adeyemi Olusola:	but lets try reorder...I think we should be able to pull something from it, though not sure about the heatmap thingy
00:54:26	Adeyemi Olusola:	on our own though*
01:05:38	Maria Eleni Soilemezidi:	no worries! Thank you for the presentation, Matthew! :)
01:05:39	Freya Watkins:	Thanks Matthew!
01:06:44	Maria Eleni Soilemezidi:	bye everyone, see you next week!
```
</details>


### Cohort 7

{{< video https://www.youtube.com/embed/F-vr9yl_hmQ >}}


### Cohort 8

{{< video https://www.youtube.com/embed/kWcA_ES2VtI >}}

<details>
<summary> Meeting chat log </summary>

```
00:42:29	Ahmed:	https://www.causact.com/index.html#welcome
00:42:54	Abdou:	Thanks!
```
</details>
