# Pipes

**Learning objectives:**

- What is the ` %>% ` operator and how to use it
  - `lhs` versus `rhs`
- When NOT to use the ` %>% ` operator
- Review other pipe operators that may be helpful

## Introduction

- The pipe operator comes from the `{magrittr}` package by [Stefan Milton Bache](https://stefanbache.dk/)
- If you use ***ANY*** tidyverse libraries, `{magrittr}` is already included (how to validate this statement)
- For learning humans....read the `%>%` operator as semantically: *and then*

```{r, Loading magrittr, eval=FALSE}
library(magrittr)
```


## Lets see it in Code!!!

The following example will highlight a code snippet **without** piping (Object Assignment and Nested) and **with** piping.

```{r, Example without Pipe, eval = FALSE}
a <- dplyr::filter(mtcars, carb > 1)
b <- dplyr::group_by(a, cyl)
c <- dplyr::summarise(b, Avg_mpg = mean(mpg))
d <- dplyr::arrange(c, desc(Avg_mpg))
print(d)
```
Above, objects `a-d` are created as "placeholders". Objects that are nothing more then memory space to hold the output of a function. The difference is, each of these can be called upon in future script.

> One of my favorite talks: [Jenny Bryant at Plotcon, 2016](https://youtu.be/4MfUCX_KpdE). Specific to the topic, start at min 15:00. The discussion is related to Dataframes and the key takeaway is wrangle your data once.

But couldn't I just nest my function?
```{r, Nested Call, eval = FALSE}
dplyr::arrange(
   dplyr::summarize(
       dplyr::group_by(
           dplyr::filter(mtcars, carb > 1),
           cyl
          ),
       Avg_mpg = mean(mpg)
      ),
   desc(Avg_mpg)
 )
```
Nesting a long function is acceptable and in some languages preferred. However, given the structure of RStudio and its plethora of packages, there is no need.

Using the Pipe
```{r, Example with Pipe, eval=FALSE}
mtcars %>%
        dplyr::filter(carb > 1) %>%
        dplyr::group_by(cyl) %>%
        dplyr::summarise(Avg_mpg = mean(mpg)) %>%
        dplyr::arrange(desc(Avg_mpg))
```
The third example not only reduces complexity....you are building a pipeline to process the `mtcars` dataset. There is no difference in computation from the first example....only, now everything is in memory.

>NOTE: There is a section in here about `pryr` that I wasn't sure the intent other then a discussion on memory use. R is intellegent and will share colums across dataframes. The material was excluded from the discussion.

## Overwritting Data....easy, right?
One method of wrangling data is to just overwrite and reassign a calculation to its original object. But, there is a problem...

1. Overwriting an object prevents clean debugging. You'll have to rerun your code over and over to see what is changing.
2. The repetition of the object being transformed implies...."what does my data look like now?"

If you want to make yourself crazy...continually read in a new copy of your data just to walk through the process over agian. This is *not* conducive to efficiency.

## The pipe makes lexical transformations
- `magrittr` reassembles the code in the pipe to a form that works by overwriting an intermediate object, see [link](https://adv-r.hadley.nz/functions.html#lexical-scoping)
- This means that the pipe won’t work for two classes of functions:
 
 1. Functions that use the current environment.
 2. Functions that use lazy evaluation.
 
## When not to use the pipe
- Your pipes are longer than (say) ten steps.
- You have multiple inputs or outputs.
- You are starting to think about a directed graph with a complex dependency structure.

## Other tools from `magrittr`
Before we begin, lets discuss `lhs` and `rhs`

- `lhs` - Left Hand Side
- `rhs` - Right Hand Side

Alternative Pipes

- **Tee** pipe: `%T>%`
  - When working with more complex pipes, it’s sometimes useful to call a function for its side-effects
    - Returns the left-hand side instead of the right-hand side
  
```{r, Tee Pipe, include=TRUE}
library(magrittr)
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()

# Using hte Tee pipe allows for seeing the data prior to plotting
rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
```

- **Exposition** pipe: `%$%`
  - It *explodes* out the variables in a data frame so that you can refer to them explicitly
 
```{r, Exposition Pipe, include = TRUE}
# Output the correlation of `disp` to `mpg`
mtcars %$%
  cor(disp, mpg)
```
 
- **Assignment** pipe: `%<>%`
  - Allows you to replace code
 
```{r, Without Assignment, include = TRUE}
mtcars <- mtcars %>% 
  transform(cyl = cyl * 2)
```

Using the Assignment Operator `%<>%`

```{r, With Assignment, include = TRUE}
mtcars %<>% transform(cyl = cyl * 2)
```

> *In my opinion, a little bit of duplication (i.e. repeating the name of the object twice) is fine in return for making assignment more explicit.* - Hadley Wickham
 