# Hierarchical data

**Learning objectives:**

- ADD ABOUT ONE THING PER SECTION DESCRIBING WHAT YOU LEARN IN THIS CHAPTER.

## SLIDE TITLE

- ADD NEW SLIDES USING ##. 
- TRY TO KEEP THE FEEL SLIDE-LIKE
  - BULLETED LISTS, NOT PARAGRAPHS

## Meeting Videos

### Cohort 7

`r knitr::include_url("https://www.youtube.com/embed/5t0xD2XsKYM")`

<details>
  <summary> Meeting chat log </summary>
```
00:04:06	Oluwafemi Oyedele:	Hi Tim, Good Evening!!!
00:04:46	Tim Newby:	Good evening - I’m struggling with a new webcam so not sure if sound/video will be working for me tonight!
00:04:49	Oluwafemi Oyedele:	Let us wait a few minute for others to join!!!
00:05:08	Oluwafemi Oyedele:	No problem!!!
00:05:22	Oluwafemi Oyedele:	You can use the chat!!!
00:11:49	Oluwafemi Oyedele:	start
00:21:08	Oluwafemi Oyedele:	https://tidyr.tidyverse.org/reference/unnest_wider.html
00:21:21	Oluwafemi Oyedele:	https://tidyr.tidyverse.org/reference/unnest_longer.html
00:52:48	Oluwafemi Oyedele:	stop
```
</details>

### Cohort 8

`r knitr::include_url("https://www.youtube.com/embed/jY5Mb82v77c")`

### Cohort 9

`r knitr::include_url("https://www.youtube.com/embed/F53PlR1rEqs")`

## Learning objectives

1. Introduce a new data structure called lists
2. Learn how to unpack lists systematically using only two new functions: `unnest_longer()` and `unnest_wider()`
3. Get acquainted with JSON

## Motivating example: Github repos

- Consider `gh_repos`. 

```{r}
library(tidyverse)
library(repurrrsive)
is.list(gh_repos)
repos <- tibble(json = gh_repos)
repos
str(repos)
```

- Now that was overwhelming! 

- A tibble with a column of lists

- Lists could be named or unnamed. 

  - The number of elements of a named list tend to be the same in every row.  
  - Named lists tend to have the same names in every row. 
  - The number of elements of an unnamed list tend to vary from row-to-row.

- To unpack an unnamed list, use `unnest_longer()` and point to the column which is a list-column.

```{r}
repos |> unnest_longer(json) 
```

- Notice another column of lists: but this time, we have named lists. 

- Use `unnest_wider()`.

```{r}
repos |> unnest_longer(json) |>
  unnest_wider(json)
```

- Give a sense of the columns.

```{r}
repos |> unnest_longer(json) |>
  unnest_wider(json) |>
  names()
```

- We can further unpack. 

```{r}
repos |> unnest_longer(json) |>
  unnest_wider(json) |>
  select(where(is.list))
repos |> unnest_longer(json) |>
  unnest_wider(json) |>
  select(where(is.list)) |>
  unnest_wider(owner)
```

- We can also answer Exercise 23.4.4, Item 1. Roughly estimate when `gh_repos` was created. Why can you only roughly estimate the date?

```{r}
repos |> unnest_longer(json) |>
  unnest_wider(json) |>
  select(created_at)
repos |> unnest_longer(json) |>
  unnest_wider(json) |>
  select(created_at) |>
  separate_wider_position(created_at, 
                          widths = c(ymd = 10, 10)) |>
  slice_max(order_by = ymd)
```

## Motivating example: Game of Thrones

```{r}
chars <- tibble(json = got_chars)
chars
chars |> 
  unnest_wider(json)
chars |> 
  unnest_wider(json) |> 
  select(id, where(is.list))
```

- Create table of titles. 

```{r}
titles <- chars |> 
  unnest_wider(json) |> 
  select(id, titles) |> 
  unnest_longer(titles) |> 
  filter(titles != "")
titles
```


- Follow the steps used for titles to create similar tables for the aliases, allegiances, books, and TV series for the Game of Thrones characters.

```{r}
aliases <- chars |> 
  unnest_wider(json) |> 
  select(id, aliases) |> 
  unnest_longer(aliases) |> 
  filter(aliases != "") 
allegiances <- chars |> 
  unnest_wider(json) |> 
  select(id, allegiances) |> 
  unnest_longer(allegiances) |> 
  filter(allegiances != "") 
books <- chars |> 
  unnest_wider(json) |> 
  select(id, books) |> 
  unnest_longer(books) |> 
  filter(books != "") 
tvSeries <- chars |> 
  unnest_wider(json) |> 
  select(id, tvSeries) |> 
  unnest_longer(tvSeries) |> 
  filter(tvSeries != "") 
```

- Joining together: "You might expect to see this data in its own table because it would be easy to join to the characters data as needed."

```{r}
characters <- chars |> 
  unnest_wider(json) |> 
  select(id, name, gender, culture, born, died, alive)
characters
characters |> 
  left_join(titles, join_by(id)) |>
  left_join(aliases, join_by(id)) |>
  left_join(allegiances, join_by(id)) |>
  left_join(books, join_by(id)) |>
  left_join(tvSeries, join_by(id))
```


-   Exercise 23.4.4 Item 4: "Explain the following code line-by-line. Why is it interesting? Why does it work for `got_chars` but might not work in general?"

```{r}
tibble(json = got_chars) |> 
  unnest_wider(json) |> 
  select(id, where(is.list)) |> 
  pivot_longer(
    where(is.list), 
    names_to = "name", 
    values_to = "value"
  ) |>  
  unnest_longer(value)
```

## JSON

- Examples of toy JSONs from Exercise 23.5.4 Item 1: "Rectangle the `df_col` and `df_row` below. They represent the two ways of encoding a data frame in JSON."

```{r}
library(jsonlite)
json_col <- parse_json('
  {
    "x": ["a", "x", "z"],
    "y": [10, null, 3]
  }
')
json_row <- parse_json('
  [
    {"x": "a", "y": 10},
    {"x": "x", "y": null},
    {"x": "z", "y": 3}
  ]
')

df_col <- tibble(json = list(json_col)) 
df_row <- tibble(json = json_row)
df_col |> 
  unnest_wider(json) |> 
  unnest_longer(x) |> 
  unnest_longer(y)
df_row |> unnest_wider(json)
``` 

- "Wilder" JSON: Try opening in a text editor and browser the Game of Thrones JSON. Below you will see a path to it. 

```{r}
got_chars_json()
```

