# Quarto formats

## Learning objectives

-   Render Quarto documents to different formats.
-   Set output options for Quarto documents in different formats.
-   Render documents in various formats from Quarto documents.
-   Render presentations in various formats from Quarto documents.
-   Create interactive web pages with Quarto.
-   Group Quarto documents together into websites and books.
-   Find information about other Quarto formats.

## Setting output type

2 ways to set qmd output format:

1. YAML header:

``` yaml
title: "Diamond sizes"
format: html
```

2. `quarto::quarto_render()`:

```{r}
#| eval: false

quarto::quarto_render("diamond-sizes.qmd", output_format = c("docx", "pdf"))
```

Note: `output_format` can have multiple values

## Output options

-   [Quarto output formats](https://quarto.org/docs/output-formats/all-formats.html)
-   Some options shared (`toc: true`)
-   Some format-specific (`code-fold: true` for `format: html`)
-   `default` to use defaults for format

``` yaml
format:
  html:
    toc: true
    toc_float: true
  pdf: default
  docx: default
```

-   `output_format = "all"` for all specified formats

## Documents

Document formats:

-   `pdf` = PDF with LaTeX 
-   `docx` = Microsoft Word
-   `gfm` = GitHub Flavored Markdown
-   `ipynb` = Jupyter Notebooks

## Presentations

-   `#` = new section, `##` = new slide

Presentation formats:

-   `revealjs` = HTML presentation with revealjs
-   `pptx` = PowerPoint presentation
-   `beamer` = PDF presentation with LaTeX Beamer

More at [quarto.org/docs/presentations](https://quarto.org/docs/presentations)

## Interactivity

`format: html` support interactivity

-   {htmlwidgets} = pure client-side interactivity
-   {shiny} = complex interactivity with server-side processing
-   {shinylive} (not in book) = complex interactivity without server

## htmlwidgets

-   Handles HTML & JavaScript for you
-   Many packages provide {htmlwidgets}:
    -   [{leaflet}](https://rstudio.github.io/leaflet/) = maps
    -   [{dygraphs}](https://rstudio.github.io/dygraphs) = time series visualizations
    -   [{DT}](https://rstudio.github.io/DT/) = tables
    -   [{threejs}](https://bwlewis.github.io/rthreejs) = 3d plots
    -   [{DiagrammeR}](https://rich-iannone.github.io/DiagrammeR) = flow charts, node-link diagrams, etc
    -   More at [htmlwidgets.org](https://www.htmlwidgets.org)

## Shiny

``` yaml
title: "Shiny Web App"
format: html
server: shiny
```

```{r}
#| eval: false

library(shiny)

textInput("name", "What is your name?")
numericInput("age", "How old are you?", NA, min = 0, max = 150)
```

-   Code chunk with `#| context: server` for server-side code

## ShinyLive

-   Uses [WebR](https://docs.r-wasm.org/webr/latest/)
-   No server side
-   Can't (yet?) communicate with internet
-   Can't hide secrets from user

[quarto-ext.github.io/shinylive](https://quarto-ext.github.io/shinylive/)

## Websites and books

Quarto projects can group `.qmd`s into websites or books

-   Put `.qmd` files in single directory (confirm)
-   `index.qmd` = home page
-   `_quarto.yml` specifies structure

``` yaml
project:
  type: book
```

-   More at [quarto.org](https://quarto.org):
    -   [/docs/websites](https://quarto.org/docs/websites)
    -   [/docs/books](https://quarto.org/docs/books)

## Other formats

-   [Quarto Journal Templates](https://quarto.org/docs/journals/templates.html)
-   [More at quarto.org](https://quarto.org/docs/output-formats/all-formats.html)

## Resources

-   [quarto.pub](https://quartopub.com/): Quarto publishing platform
-   [github.com/mcanouil/awesome-quarto](https://github.com/mcanouil/awesome-quarto): Awesome Quarto (Quarto talks, tools, etc) 
-   [quarto.org](https://quarto.org/): Quarto documentation
-   [github.com/quarto-dev](https://github.com/quarto-dev)

## Meeting Videos

### Cohort 5

{{< video https://www.youtube.com/embed/2MGp74nb2ig >}}

<details>
  <summary> Meeting chat log </summary>
  
```
00:10:00	Jon Harmon (jonthegeek):	bestbook.cool
00:11:26	Jon Harmon (jonthegeek):	https://podcast.bestbook.cool/
00:28:01	Jon Harmon (jonthegeek):	happygitwithr.com
00:30:59	Jon Harmon (jonthegeek):	Remind @Sandra Muroy to share the dlab GitHub stuff when this posts to our channel. There, I think that will tag you in the post!
00:32:44	Jon Harmon (jonthegeek):	xaringan
00:36:45	Jon Harmon (jonthegeek):	lol, it's just called {shinydashboard}
00:48:24	Becki R. (she/her):	I definitely need help with that!
00:53:08	Jon Harmon (jonthegeek):	It's been a while since we learned that!
2021-08-21: Chapter 6: Workflow: scripts: Ryan Metcalf
01:01:01	Jon Harmon (jonthegeek):	{pak} = updated package tracker
01:01:48	Jon Harmon (jonthegeek):	{targets}
01:03:00	Federica Gazzelloni:	library(pkgsnap)
01:04:15	Jon Harmon (jonthegeek):	devtools::install_version()
01:05:00	Jon Harmon (jonthegeek):	?remotes::install_version()
01:05:15	Jon Harmon (jonthegeek):	install_version("devtools", ">= 1.12.0, < 1.14")
01:07:07	Jon Harmon (jonthegeek):	https://github.com/MangoTheCat/pkgsnap
01:11:32	Jon Harmon (jonthegeek):	https://docs.ropensci.org/targets/
01:11:46	Jon Harmon (jonthegeek):	pkgdown
01:12:41	Jon Harmon (jonthegeek):	https://rsample.tidymodels.org/
01:13:21	Jon Harmon (jonthegeek):	https://pkgdown.r-lib.org/
01:13:32	Sandra Muroy:	https://github.com/dlab-berkeley/Bash-Git
```
</details>

### Cohort 6

{{< video https://www.youtube.com/embed/qQrnNef9fkM >}}

<details>
  <summary> Meeting chat log </summary>
  
```
00:09:51	Marielena Soilemezidi:	http://www.thefunctionalart.com/p/instructors-guide.html
https://www.dropbox.com/s/gyqyz3hei7lhhmt/R_RESOURCES.txt?dl=0
https://exts.ggplot2.tidyverse.org/gallery/
00:26:20	Daniel:	underscore
00:26:32	Daniel:	It is called an “underscore”
00:42:17	Daniel:	I can’t see the screen anymore
00:43:00	Daniel:	All good now
00:44:57	Marielena Soilemezidi:	https://github.com/jrnold/ggthemes)
00:55:42	Daniel:	https://stackoverflow.com/questions/23957278/how-to-add-table-of-contents-in-rmarkdown
01:03:59	Daniel:	It’s fine, sorry my internet went off
01:04:17	Marielena Soilemezidi:	http://shiny.rstudio.com/.
01:05:04	Marielena Soilemezidi:	http://rmarkdown.rstudio.com/rmarkdown_websites.html.
01:07:21	Marielena Soilemezidi:	http://colinpurrington.com/tips/lab-notebooks
01:10:02	Marielena Soilemezidi:	https://quarto.org/
```
</details>


### Cohort 7

{{< video https://www.youtube.com/embed/vkjYlBK9-9c >}}

<details>
<summary> Meeting chat log </summary>

```
00:08:34	Oluwafemi Oyedele:	We will start in 5 minute time!!!
00:14:37	Oluwafemi Oyedele:	start
00:52:30	Oluwafemi Oyedele:	https://github.com/quarto-dev/quarto-cli/tree/main/src/resources/formats/revealjs/themes
01:09:33	Oluwafemi Oyedele:	https://quarto.org/docs/journals/templates.html
01:10:52	Oluwafemi Oyedele:	stop
```
</details>


### Cohort 8

{{< video https://www.youtube.com/embed/7kW7DEWPVZ8 >}}

<details>
<summary> Meeting chat log </summary>

```
00:01:59	Shamsuddeen Muhammad:	start
00:10:33	Ahmed Mamdouh:	Im here
00:10:47	Ahmed Mamdouh:	The internet is so bad sorry
00:12:18	Ahmed Mamdouh:	Oh interesting
00:18:39	Ahmed Mamdouh:	It like packaging all dependencies
00:25:58	Ahmed Mamdouh:	Renv is like venv in python
00:29:57	Ahmed Mamdouh:	Yup
00:31:32	Ahmed Mamdouh:	Awesome
00:32:25	Shamsuddeen Muhammad:	stop
00:32:27	Ahmed Mamdouh:	Thanks alot man
00:32:50	Ahmed Mamdouh:	Bye
```
</details>
