# Quarto formats

**Learning objectives:**

- Give a brief overview of some options for communicating your results with Quarto, from static and interactive documents to presentations to websites and books.


## Setting output type

There are two ways to set the output of a document:


1. Modify the YAML header:
```{r eval=FALSE}
title: "Diamond sizes"
format: html
```

2. Call quarto::quarto_render() by hand:

```{r eval=FALSE}

quarto::quarto_render("diamond-sizes.qmd", output_format = c("docx", "pdf"))
```

Note: You can  produce multiple types of output since the `output_format` argument can take a list of values

## Output options

- You can find the complete list of Quarto output formats at <https://quarto.org/docs/output-formats/all-formats.html>
- Many formats share some output options, but others have options that are format specific.
  - Ex. including a table of content vs. code folding
- To render to all formats specified in the YAML of a document, you can use `output_format = "all"`.

```{r eval=FALSE}
quarto::quarto_render("diamond-sizes.qmd", output_format = "all")
```

## Documents

- There are several basic variations off HTML for generating different types of documents. To name a few, examples are:
  - `pdf` makes a PDF with LaTeX 

  - `docx` for Microsoft Word (.docx) documents.

  - `gfm` for a GitHub Flavored Markdown (.md) document.

  - `ipynb` for Jupyter Notebooks (.ipynb).


```{r eval=FALSE}
format:
  html:
    toc: true #include table of contents
    toc_float: true #make table of contents floating 
  pdf: default
  docx: default
```

Note: use special syntax (i.e. `pdf: default`) if you don’t want to override any default options.

## Presentations

- When using Quarto to produce presentations, you get less visual control than with a tool like PowerPoint, but the upside is you can automatically insert the results of your R code into your presentation.
- Presentations work by dividing your content into slides, with a new slide beginning at each second (`##`) level header.
  - First (`#`) level headers indicate the beginning of a new section with a section title slide.
- Quarto supports a variety of presentation formats, including:

  - `revealjs` - HTML presentation with revealjs
  
  - `pptx` - PowerPoint presentation
  
  - `beamer` - PDF presentation with LaTeX Beamer.

- You can read more about creating presentations with Quarto at <https://quarto.org/docs/presentations>.

## Interactivity

- HTML documents created with Quarto can contain interactive components as well. 
- Two example options for including interactivity in your Quarto documents are htmlwidgets and Shiny.
  - **htmlwidgets** are R functions that produce interactive HTML visualizations
  - `Shiny` is a package that allows you to create interactivity using R code

## htmlwidgets

- htmlwidgets provide client-side interactivity, meaning all the interactivity happens in the browser, independently of R. 
- You don’t need to know anything about HTML or JavaScript to use them because all the details are wrapped inside the package.
- There are many packages that provide htmlwidgets, including:
  - [`leaflet`](https://rstudio.github.io/leaflet/) for interactive map
  - [`dygraphs`](https://rstudio.github.io/dygraphs) for interactive time series visualizations.
  
  - [`DT`](https://rstudio.github.io/DT/) for interactive tables.
  
  - [`threejs`](https://bwlewis.github.io/rthreejs) for interactive 3d plots.
  
  - [`DiagrammeR`](https://rich-iannone.github.io/DiagrammeR) for diagrams (like flow charts and simple node-link diagrams).

- To learn more about htmlwidgets and see a complete list of packages that provide them visit <https://www.htmlwidgets.org>.

## Shiny

- Shiny interactions occur on the server-side, meaning you can write interactive apps without knowing JavaScript, but you need a server to run them on. 

- To call Shiny code from a Quarto document, add `server: shiny` to the YAML header:
```{r eval=FALSE}
title: "Shiny Web App"
format: html
server: shiny
```

- Then you can use the “input” functions to add interactive components to the document:

```{r eval=FALSE}
library(shiny)

textInput("name", "What is your name?")
numericInput("age", "How old are you?", NA, min = 0, max = 150)
```

- And you also need a code chunk with chunk option `context: server` which contains the code that needs to run in a Shiny server.

## Websites and books

- With a bit of additional infrastructure, you can use Quarto to generate a complete website or book:

  - Put your `.qmd` files in a single directory. `index.qmd` will become the home page.
  
  - Add a YAML file named `_quarto.yml` that provides the navigation for the site. In this file, set the project type to either `book` or `website`, e.g.:
  
```{r eval=FALSE}
project:
  type: book
```

- Read more at <https://quarto.org/docs/websites> about Quarto websites and <https://quarto.org/docs/books> about books.

- Note: [R4DS 2E](r4ds.hadley.nz) is itself a quarto book. You can find the code to render it on your own here: <https://github.com/hadley/r4ds/>. The [Quarto website](https://quarto.org/) itself was made with Quarto. You can find the code here: <https://github.com/quarto-dev/quarto-web>.

## Other formats

- Quarto offers even more output formats. For example:

  - You can write journal articles using Quarto Journal Templates: <https://quarto.org/docs/journals/templates.html>.
  
  - You can output Quarto documents to Jupyter Notebooks with `format: ipynb`: <https://quarto.org/docs/reference/formats/ipynb.html>.

- See <https://quarto.org/docs/output-formats/all-formats.html> for a list of even more formats.

## Resources

- [Quarto Pub](https://quartopub.com/): Create documents, websites, presentations, books, and blogs in Quarto, then securely publish them to the web with the Quarto CLI, the easiest way to publish and share on the web.
- [Awesome Quarto](https://github.com/mcanouil/awesome-quarto): A curated list of Quarto talks, tools, examples & articles! 
- Quarto documentation: <https://quarto.org/>
- Quarto GitHub repo: <https://github.com/quarto-dev>

## Meeting Videos

### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/2MGp74nb2ig")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:10:00	Jon Harmon (jonthegeek):	bestbook.cool
00:11:26	Jon Harmon (jonthegeek):	https://podcast.bestbook.cool/
00:28:01	Jon Harmon (jonthegeek):	happygitwithr.com
00:30:59	Jon Harmon (jonthegeek):	Remind @Sandra Muroy to share the dlab GitHub stuff when this posts to our channel. There, I think that will tag you in the post!
00:32:44	Jon Harmon (jonthegeek):	xaringan
00:36:45	Jon Harmon (jonthegeek):	lol, it's just called {shinydashboard}
00:48:24	Becki R. (she/her):	I definitely need help with that!
00:53:08	Jon Harmon (jonthegeek):	It's been a while since we learned that!
2021-08-21: Chapter 6: Workflow: scripts: Ryan Metcalf
01:01:01	Jon Harmon (jonthegeek):	{pak} = updated package tracker
01:01:48	Jon Harmon (jonthegeek):	{targets}
01:03:00	Federica Gazzelloni:	library(pkgsnap)
01:04:15	Jon Harmon (jonthegeek):	devtools::install_version()
01:05:00	Jon Harmon (jonthegeek):	?remotes::install_version()
01:05:15	Jon Harmon (jonthegeek):	install_version("devtools", ">= 1.12.0, < 1.14")
01:07:07	Jon Harmon (jonthegeek):	https://github.com/MangoTheCat/pkgsnap
01:11:32	Jon Harmon (jonthegeek):	https://docs.ropensci.org/targets/
01:11:46	Jon Harmon (jonthegeek):	pkgdown
01:12:41	Jon Harmon (jonthegeek):	https://rsample.tidymodels.org/
01:13:21	Jon Harmon (jonthegeek):	https://pkgdown.r-lib.org/
01:13:32	Sandra Muroy:	https://github.com/dlab-berkeley/Bash-Git
```
</details>

### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/qQrnNef9fkM")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:09:51	Marielena Soilemezidi:	http://www.thefunctionalart.com/p/instructors-guide.html
https://www.dropbox.com/s/gyqyz3hei7lhhmt/R_RESOURCES.txt?dl=0
https://exts.ggplot2.tidyverse.org/gallery/
00:26:20	Daniel:	underscore
00:26:32	Daniel:	It is called an “underscore”
00:42:17	Daniel:	I can’t see the screen anymore
00:43:00	Daniel:	All good now
00:44:57	Marielena Soilemezidi:	https://github.com/jrnold/ggthemes)
00:55:42	Daniel:	https://stackoverflow.com/questions/23957278/how-to-add-table-of-contents-in-rmarkdown
01:03:59	Daniel:	It’s fine, sorry my internet went off
01:04:17	Marielena Soilemezidi:	http://shiny.rstudio.com/.
01:05:04	Marielena Soilemezidi:	http://rmarkdown.rstudio.com/rmarkdown_websites.html.
01:07:21	Marielena Soilemezidi:	http://colinpurrington.com/tips/lab-notebooks
01:10:02	Marielena Soilemezidi:	https://quarto.org/
```
</details>


### Cohort 7

`r knitr::include_url("https://www.youtube.com/embed/vkjYlBK9-9c")`

<details>
<summary> Meeting chat log </summary>

```
00:08:34	Oluwafemi Oyedele:	We will start in 5 minute time!!!
00:14:37	Oluwafemi Oyedele:	start
00:52:30	Oluwafemi Oyedele:	https://github.com/quarto-dev/quarto-cli/tree/main/src/resources/formats/revealjs/themes
01:09:33	Oluwafemi Oyedele:	https://quarto.org/docs/journals/templates.html
01:10:52	Oluwafemi Oyedele:	stop
```
</details>


### Cohort 8

`r knitr::include_url("https://www.youtube.com/embed/7kW7DEWPVZ8")`

<details>
<summary> Meeting chat log </summary>

```
00:01:59	Shamsuddeen Muhammad:	start
00:10:33	Ahmed Mamdouh:	Im here
00:10:47	Ahmed Mamdouh:	The internet is so bad sorry
00:12:18	Ahmed Mamdouh:	Oh interesting
00:18:39	Ahmed Mamdouh:	It like packaging all dependencies
00:25:58	Ahmed Mamdouh:	Renv is like venv in python
00:29:57	Ahmed Mamdouh:	Yup
00:31:32	Ahmed Mamdouh:	Awesome
00:32:25	Shamsuddeen Muhammad:	stop
00:32:27	Ahmed Mamdouh:	Thanks alot man
00:32:50	Ahmed Mamdouh:	Bye
```
</details>
