# Many models {-}

**Learning objectives:**

-   Create nested data frames to organize data by groups.
-   Create list-columns to generate new data in an organized structure.
-   Simplify list-columns to manipulate the data they contain.

## Introduction

-   The main purpose of this section.....is to tidy your data and how model summaries can help us pick out outliers and unusual trends in our data
-   As your experience grows with Exploratory Data Analysis (EDA) you will find your models grow as well

> This chapter is somewhat aspirational: if this book is your first introduction to R, this chapter is likely to be a struggle. It requires you to have deeply internalised ideas about modelling, data structures, and iteration. So don't worry if you don't get it --- just put this chapter aside for a few months, and come back when you want to stretch your brain.

-   We are not going to discuss Model Building (even though it is the name of the Chapter)
-   We are going to push through to the end....(but please feel welcome to ask questions!)

### Prerequisites

We will use both `tidyverse` and `modelr` from here.

```{r 99-25-Load Packages}
library(modelr)
library(tidyverse)
```

## gapminder

Where do I begin to express the awesomness of Hans Rosling (27 July 1948 -- 7 February 2017). I HIGHLY encourage you to watch the youtube video!

[Hans Rosling's 200 Countries, 200 Years, 4 Minutes - The Joy of Stats - BBC Four](https://www.youtube.com/watch?v=jbkSRLYSojo%22)

Furthermore, we can thank Jenny Bryan for authoring the gapminder package!

```{r 99-25-Load Gapminder}
library(gapminder)
```

To gain some insight to the data, we ask the question: "How does life expectancy (lifeExp) change over time (year) for each country (country)?"

```{r 99-25-Plot Gapmider}
gapminder %>%
  ggplot(aes(year, lifeExp, group = country)) +
    geom_line(alpha = 1/3)

```

At first glance, it appears life expectancy is increasing...but not for all countries.

To make it easier to view, we fit a model with a linear trend. The model captures steady growth over time, and the residuals will show what's left.

```{r 99-25-Linear Trend}
nz <- filter(gapminder, country == "New Zealand")
nz %>%
  ggplot(aes(year, lifeExp)) +
  geom_line() +
  ggtitle("Full data = ")

nz_mod <- lm(lifeExp ~ year, data = nz)
nz %>%
  add_predictions(nz_mod) %>%
  ggplot(aes(year, pred)) +
  geom_line() +
  ggtitle("Linear trend + ")

nz %>%
  add_residuals(nz_mod) %>%
  ggplot(aes(year, resid)) +
  geom_hline(yintercept = 0, colour = "white", linewidth = 3) +
  geom_line() +
  ggtitle("Remaining pattern")
```
This is good....but how do we do this for every country?

### Nested data

To make life easier for us (and ensure we are not copying and pasting forever), we use the `map` function from the `purrr` package.

Here, we are going to make a **nested data frame**.

```{r 99-25-Nested Data Frame}
by_country <- gapminder %>%
  group_by(country, continent) %>%
  nest()

by_country
```
This creates a data frame that has one row per group (per country), and a rather unusual column:`data`. `data` is a list of data frames (or tibbles, to be precise).

>Note: Don't use the Structure function `str()` as it will be difficult to view. Instead, just view a single line of your nested dataframe.

```{r 99-25-View the Nested Dataframe}
by_country$data[[1]]

```
 >Note the difference between a standard grouped data frame and a nested data frame: in a grouped data frame, each row is an observation; in a nested data frame, each row is a group.

### List-columns

Now, lets fit some models!

```{r 99-25-Fit some Models}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}
models <- map(by_country$data, country_model)
```

Yet, this may be too costly, so instead, lets modify (be more elegant) with our data. Storing related objects in columns is a key part of the value of data frames.

```{r 99-25-Mutate}
by_country <- by_country %>%
  mutate(model = map(data, country_model))
by_country
```

This is a huge advantage: because all the related objects are stored together, you don’t need to manually keep them in sync when you filter or arrange.

```{r 99-25-Filter by Country}
by_country %>%
  filter(continent == "Europe")

by_country %>%
  arrange(continent, country)
```

***If your list of data frames and list of models were separate objects, you have to remember that whenever you re-order or subset one vector, you need to re-order or subset all the others in order to keep them in sync. If you forget, your code will continue to work, but it will give the wrong answer!***

### Unnesting
Previously, we were working with only one country. Now, lets compute the residuals of ALL countries.

```{r 99-25-Compute Residuals}
by_country <- by_country %>%
  mutate(
    resids = map2(data, model, add_residuals)
  )
by_country
```
By now, you should ask yourself: "How can I plot a bunch of dataframes?". This isn't required. We have to *un-nest*.

```{r 99-25-Un-Nest}
resids <- unnest(by_country, resids)
resids
```
Now, we can plot ALL the residuals.

```{r 99-25-Plot Residuals}
resids %>%
  ggplot(aes(year, resid)) +
    geom_line(aes(group = country), alpha = 1 / 3) +
    geom_smooth(se = FALSE)
```
We can now plot each country as a facet.

```{r 99-25-Facet by Country}
resids %>%
  ggplot(aes(year, resid, group = country)) +
    geom_line(alpha = 1 / 3) +
    facet_wrap(~continent)
```
We can note that we still have very large residuals, namely in Africa suggesting our model isn't fitting very well.

### Model quality

Instead of looking at the residuals from the model, we could look at some general measurements of model quality.

We’ll use `broom::glance()` to extract some model quality metrics.

```{r 99-25-Broom::Glance}
broom::glance(nz_mod)
```
We can use mutate() and unnest() to create a data frame with a row for each country.

```{r Broom:Glance Un-Nested}
by_country %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance)
```
This isn’t quite the output we want, because it still includes all the list columns. This is default behaviour when `unnest()` works on single row data frames. To suppress these columns we use `.drop = TRUE`

>Note, `.drop=TRUE` has been depricated.

```{r 99-25-"Un-Nest w/ .drop=TRUE"}
glance <- by_country %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance, .drop = TRUE)
glance
```

Now, we can look for countries that don't fit our model well.

```{r 99-25-Sort by Residuals}
glance %>%
  arrange(r.squared)
```
The worst fitting models seem to be in Africa. We'll add `geom_jitter()` to make it more apparent.

We can also plot the particular bad $$R^2$$ and plot the data.

```{r 99-25-Geom_Jitter plus bad R2}
glance %>%
  ggplot(aes(continent, r.squared)) +
    geom_jitter(width = 0.5)

bad_fit <- filter(glance, r.squared < 0.25)

gapminder %>%
  semi_join(bad_fit, by = "country") %>%
  ggplot(aes(year, lifeExp, colour = country)) +
    geom_line()
```
The relation of this visual is the tragidies of HIV/AIDs epidemic and the Rwanda genocide.

## List-columns

- List-Columns are implicit in the definition of the data frame: a data frame is a named list of equal length vectors.
- Base-R doesn't make it easy to create list-columns, and `data.frame()` treats a list as a list of columns.

```{r 99-25-List of Columns}
data.frame(x = list(1:3, 3:5))
```

- You can prevent `data.frame()` from treating a lists of lists by adding `I()` to the argument. However, this doesn't print well.

`I()` stands for **Inhibit Interpretation/Conversion of Objects**: Change the class of an object to indicate that is should be treated *as is*.

```{r 99-25-using I}
data.frame(
  x = I(list(1:3, 3:5)),
  y = c("1, 2", "3, 4, 5")
)
```

- Tibble alleviates this problem by being lazier (`tibble()` doesn’t modify its inputs) and by providing a better print method.

>Note where the quotes are placed

```{r 99-25-List with Tibble}
tibble(
  x = list(1:3, 3:5),
  y = c("1, 2", "3, 4, 5")
)
```
- `tribble()` can automatically work out that you need a list.

```{r 99-25-Tribble Example}
tribble(
   ~x, ~y,
  1:3, "1, 2",
  3:5, "3, 4, 5"
)
```
- List-columns are often most useful as intermediate data structure.
- Advantage of keeping related items together in a data frame is worth a little hassle.

There are three parts of an effective list-column pipeline:

1. You create the list-column using one of: `nest()`, `summarise() + list()`, or `mutate() + a map function`
2. You create other intermediate list-columns by transforming existing list columns with `map()`, `map2()`, or `pmap()`.
3. You simplify the list-column back down to a data frame or atomic vector.

## Creating list-columns

Typically, you won’t create list-columns with `tibble()`. Instead, you’ll create them from regular columns, using one of three methods:

1. With `tidyr::nest()` to convert a grouped data frame into a nested data frame where you have list-column of data frames.
2. With `mutate()` and vectorised functions that return a list.
3. With `summarise()` and summary functions that return multiple results.

>Alternatively, you might create them from a named list, using `tibble::enframe()`

When creating list-columns, make sure they are homogeneous.

### With nesting

- `nest()` creates a nested data frame, meaning, each row is a *meta-observation*.

- When applied to a group data frame, `nest()` keeps the grouping columns *as is*.

```{r 99-25-Gapminder Grouped}
gapminder %>%
  group_by(country, continent) %>%
  nest()
```
- You can also use it on an un-grouped data frame, specifying which columns you want to nest.

```{r 99-25-Specify Columns}
gapminder %>%
  nest(data = c(year:gdpPercap))
```
### From vectorised functions

- If you use `stringr::str_split() + mutate()` you get a list-column.

> Again, note where the quotes are placed.

```{r 99-25-Vectorized Functions}
df <- tribble(
  ~x1,
  "a,b,c",
  "d,e,f,g"
)

df %>%
  mutate(x2 = stringr::str_split(x1, ","))
```
- And now `unnest()` knows how to handle these list of vectors.

```{r 99-25-Un-Nested list of vectors}
df %>%
  mutate(x2 = stringr::str_split(x1, ",")) %>%
  unnest(x2)
```

>If you find yourself using this pattern a lot, make sure to check out `tidyr::separate_rows()` which is a wrapper around this common pattern.

Another example uses `map(), `map2()`, and pmap()`. We could re-write [Invoking different functions](https://r4ds.had.co.nz/iteration.html#invoking-different-functions) and rewrite it to use `mutate()`.

Previous Example Code:
```{r 99-25-Chapter 21 Code Example}
sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = 1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)
sim %>%
  mutate(sim = invoke_map(f, params, n = 10))
```

Refactored Code using `mutate()`

```{r 99-25-Refactored using Mutate}
sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = 1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)

sim %>%
  mutate(sims = invoke_map(f, params, n = 10))
```
> I donm't understand what is being expressed here.....the two code snippets are identical, except in chapter 25, the name is `sims` instead. Thoughts?

### From multivalued summaries

One restriction of `summarise()` is it only works with summary functions that return a single value. Implying, you can't use it with functions like `quantile()` that return a vector of arbitrary length.

```{r 99-25-mtcars with summarise}
mtcars %>%
  group_by(cyl) %>%
  summarise(q = quantile(mpg))
```

You can however, wrap the result in a list! This obeys the contract of summarise(), because each summary is now a list (a vector) of length 1.

```{r 99-25-Summarise with list}
mtcars %>%
  group_by(cyl) %>%
  summarise(q = list(quantile(mpg)))
```
To make useful results with `unnest90`, you'll also ned to capture probabilities.

```{r 99-25-Capture Probabilities}
probs <- c(0.01, 0.25, 0.5, 0.75, 0.99)
mtcars %>%
  group_by(cyl) %>%
  summarise(p = list(probs), q = list(quantile(mpg, probs))) %>%
  unnest(c(p, q))
```
### From a named list

What do you do if you want to iterate over both the contents of a list and its elements?

- Make a data frame with one column containing the elements and another column containing the list!

You can use `tibble::enframe()`.

```{r 99-25-Tibble Enframe}
x <- list(
  a = 1:5,
  b = 3:4,
  c = 5:6
)

df <- enframe(x)
df
```

Now, if we want to iterate over names and values in parrallel, we can use `map2()`.

```{r 99-25-map2}
df %>%
  mutate(
    smry = map2_chr(name, value, ~ stringr::str_c(.x, ": ", .y[1]))
  )
```

## Simplifying list-columns

To simplify the list-column back to a regular column (an atomic vector) or set of columns:

1. If you want a single value, use `mutate()` with `map_lgl()`, `map_int()`, `map_dbl()`, and `map_chr()` to create an atomic vector.
2. If you want many values, use `unnest()` to convert list-columns back to regular columns, repeating the rows as many times as necessary.

### List to vector

You can always summarise an object with its type and length, so this code will work regardless of what sort of list-column you have.

```{r 99-25-List Column}
df <- tribble(
  ~x,
  letters[1:5],
  1:3,
  runif(5)
)

df %>% mutate(
  type = map_chr(x, typeof),
  length = map_int(x, length)
)
```

Although being the same basic information you get from the default `tbl` print method, you can now use if for filtering.

Don’t forget about the `map_*()` shortcuts - you can use `map_chr(x, "apple")` to extract the string stored in apple for each element of x.

Use the `.null` argument to provide a value to use if the element is missing (instead of returning NULL).

```{r 99-25-List with Null}
df <- tribble(
  ~x,
  list(a = 1, b = 2),
  list(a = 2, c = 4)
)
df %>% mutate(
  a = map_dbl(x, "a"),
  b = map_dbl(x, "b", .null = NA_real_)
)
```

### Unnesting

`unnest()` works by repeating the regular columns once for each element of the list-column.

```{r 99-25-Unnest with Tibble}
tibble(x = 1:2, y = list(1:4, 1)) %>% unnest(y)
```

You cannot simultaneously unnest two columns that contain different number of elements.

```{r 99-25-Unnest Error}
# Ok, because y and z have the same number of elements in
# every row
df1 <- tribble(
  ~x, ~y,           ~z,
   1, c("a", "b"), 1:2,
   2, "c",           3
)
df1
df1 %>% unnest(c(y, z))

# Doesn't work because y and z have different number of elements
df2 <- tribble(
  ~x, ~y,           ~z,
   1, "a",         1:2,  
   2, c("b", "c"),   3
)
df2

df2 %>% unnest(c(y, z))
```

## Making tidy data with broom
The broom package provides three general tools for turning models into tidy data frames:

1. `broom::glance(model)` returns a row for each model. Each column gives a model summary: either a measure of model quality, or complexity, or a combination of the two.

2. `broom::tidy(model)` returns a row for each coefficient in the model. Each column gives information about the estimate or its variability.

3. `broom::augment(model, data)` returns a row for each row in data, adding extra values like residuals, and influence statistics.


***Any Questions?***

## Meeting Videos

### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/lO_qtXM8OH8")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:09:25	Jon Harmon (jonthegeek):	dslc.io/bookclubber
01:07:14	Jon Harmon (jonthegeek):	ggrepel
```
</details>

`r knitr::include_url("https://www.youtube.com/embed/d5FPrn4vTzc")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:10:13	Njoki Njuki Lucy:	It will be nice to see familiar faces :)
00:36:21	Jon Harmon (jonthegeek):	sim <- tribble(
  ~params,
  list(min = -1, max = 1),
  list(min = -1, max = 2),
  list(min = -1, max = 3)
)

set.seed(424242)
use_purrr <- sim %>% 
  mutate(
    sims = map(params, runif)
  )

set.seed(424242)
use_rowwise <- sim %>%
  rowwise() %>% 
  mutate(
    sims = list(runif(params))
  ) %>% 
  ungroup()

identical(use_purrr, use_rowwise)
00:37:40	Jon Harmon (jonthegeek):	> use_purrr
# A tibble: 3 x 2
  params           sims     
  <list>           <list>   
1 <named list [2]> <dbl [2]>
2 <named list [2]> <dbl [2]>
3 <named list [2]> <dbl [2]>
00:52:42	Jon Harmon (jonthegeek):	> tibble(x = 1:2, y = list(1:4, 1))
# A tibble: 2 x 2
      x y        
  <int> <list>   
1     1 <int [4]>
2     2 <dbl [1]>
00:57:04	Jon Harmon (jonthegeek):	> df2 %>% unnest(c(y))
# A tibble: 3 x 3
      x y     z        
  <dbl> <chr> <list>   
1     1 a     <int [2]>
2     2 b     <dbl [1]>
3     2 c     <dbl [1]>
00:57:23	Jon Harmon (jonthegeek):	> df2 %>% unnest(y) %>% unnest(z)
# A tibble: 4 x 3
      x y         z
  <dbl> <chr> <dbl>
1     1 a         1
2     1 a         2
3     2 b         3
4     2 c         3
01:11:05	Njoki Njuki Lucy:	that's what I have been doing - doing the surgery😄
01:11:33	Njoki Njuki Lucy:	:)
```
</details>


### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/v1OawROf1ZQ")`

`r knitr::include_url("https://www.youtube.com/embed/PX7y2jlRYuU")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:14:37	Marielena Soilemezidi:	they're called quotes/quote marks!
00:14:43	Marielena Soilemezidi:	I also got a brain freeze xD
00:42:28	Daniel:	https://github.com/tidyverse/broom
```
</details>
